---
title: "BMC_Bio_SaimiriMS1_2026"
author: "Sharmi Sen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup}
#Squirrel monkey 16S analyses: data visualization and models
rm(list = ls())

#Load libraries
invisible(lapply(c("phyloseq", "tidyverse","tidyr", "vegan", "ggplot2", "MuMIn", "microbiome", "ggpubr", "corrplot", "RColorBrewer", "sjPlot", "patchwork", "viridis", "performance", "see", "mediation", "mgcv", "lme4", "e1071", "biomformat", "cowplot", "reshape2", "survival", "compositions", "survminer", "data.table", "bbmle", "ggsci", "maaslin3", "limma", "pheatmap", "stringr", "AICcmodavg", "mice"), library, character.only = TRUE))
          
'%!in%' <- function(x,y)!('%in%'(x,y)) 
setwd("~/Desktop/POSTDOC PROJECTS/NORTHWESTERN/Squirrel Monkey Project")
```

#reading in files
```{r}
#16S data 
metadata <- readRDS("saimiri_shotgun_metadata_updated_newloc_Aug42026.rds")
ps       <- readRDS("sbol16S_ps_without_contam_before_filtering.rds") #this has negative controls but no features identified as contaminants and no features that are only present in one sample
ps       <- prune_samples(sample_sums(ps) > 0, ps) #remove taxa that are in 0 samples
ps       <- subset_samples(ps, Sample_or_Control == "Sample") #only keep samples
ps_rel   <- transform_sample_counts(ps, function(x) x / sum(x))
ps_rel #11163 taxa

greg_data <- readxl::read_excel("gkw edited _MDAnderson_SaimiriMetadata_Mar062026.xlsx")
#shotgun data with otu and SCFA
shotgun_metadata <- readRDS("shotgun_metadata_with_otudiv_funcdiv_numreads_newlocJun82026.rds") #this has otu div (alpha beta) and functional pathway div (alpha div index)
head(shotgun_metadata)
colnames(shotgun_metadata)[12] <- "Housing type"
colnames(shotgun_metadata)[16] <- "Enclosure size"
##Read in data meabolomic matrix for untargeted metabolomites data
feature_table   <- read.csv("old_stuff/sbol_metabolomics_feature_table.csv", row.names = 1, check.names = FALSE)
feature_table   <- feature_table[1:128,]
metab_metadata  <- read.csv("old_stuff/metadata_for_metabolomics.csv", row.names = 1, check.names = FALSE)
annotations     <- read.csv("old_stuff/sbol_metabolite_list.csv", row.names = 1, check.names = FALSE)
head(annotations)
#Convert to numeric
feature_matrix <- as.matrix(
  sapply(feature_table, function(col) {
    as.numeric(as.character(unlist(col)))
  }))

rownames(feature_matrix) <- rownames(feature_table)
metab_present            <- colMeans(!is.na(feature_matrix))

feature_matrix <- feature_matrix[, metab_present > 0.50]
#which are not present in shotgun data 
names_for_adding <- metab_metadata$Animal[match(rownames(feature_matrix), metab_metadata$normalized_name)]
rownames(feature_matrix) <- shotgun_metadata$Sample[match(names_for_adding, shotgun_metadata$Animal)]
#remove NA rownames
feature_matrix <- feature_matrix[which(!is.na(rownames(feature_matrix))),] 
#Remove zero-variance columns
feature_matrix_filtered <- feature_matrix[, apply(feature_matrix, 2, sd, na.rm = TRUE) > 0] #this doesn't change anything

metab_scaled <- scale(feature_matrix_filtered)
metab_dist   <- dist(metab_scaled, method = "euclidean")

#any correlations for the 16S div metrics and number of reads - should not be the case as they have been rarefied
cor.test(log(shotgun_metadata$asv_reads), shotgun_metadata$otu_shannon)
cor.test(log(shotgun_metadata$asv_reads), shotgun_metadata$otu_faith)
cor.test(log(shotgun_metadata$asv_reads), shotgun_metadata$otu_richness)

cor.test(log(shotgun_metadata$asv_reads), shotgun_metadata$otu_faith)
cor.test(log(shotgun_metadata$asv_reads), shotgun_metadata$otu_richness)
summary(lm(log(asv_reads) ~ Species, data = shotgun_metadata))

cor.test(log(shotgun_metadata$asv_reads), shotgun_metadata$otu_PC1)
cor.test(log(shotgun_metadata$asv_reads), shotgun_metadata$otu_PC2)
cor.test(log(shotgun_metadata$asv_reads), shotgun_metadata$otu_PC3)

cor.test(log(shotgun_metadata$num_reads), shotgun_metadata$shannon)
cor.test(log(shotgun_metadata$num_reads), shotgun_metadata$simpson)
cor.test(log(shotgun_metadata$num_reads), shotgun_metadata$observed_pathways)

boliv_color <- c("#2A6EBBFF")
set_color  <- c("#2A6EBBFF", "#F0AB00")
```

```{r}
#mapping pathways 
pathway_name_map <- c(
  "ANAGLYCOLYSIS-PWY" = "Anaerobic glycolysis",
  "ANAEROFRUCAT-PWY" = "Anaerobic fructose degradation",
  "ARGDEG-PWY" = "Arginine degradation I (arginine succinyltransferase pathway)",
  "ASPASN-PWY" = "L-aspartate and L-asparagine biosynthesis",
  "CALVIN-PWY" = "Calvin-Benson-Bassham cycle",
  "COA-PWY" = "Coenzyme A biosynthesis I",
  "COBALSYN-PWY" = "Adenosylcobalamin (coenzyme B12) biosynthesis",
  "DENOVOPURINE2-PWY" = "Purine nucleotide de novo biosynthesis II",
  "FOLSYN-PWY" = "Tetrahydrofolate biosynthesis",
  "GALACTUROCAT-PWY" = "D-galacturonate degradation I",
  "GLUCONEO-PWY" = "Gluconeogenesis I",
  "GLUDEG-I-PWY" = "GABA shunt",
  "GLUDEG-II-PWY" = "L-glutamate degradation VII",
  "GLYCOCAT-PWY" = "Glycolate and glyoxylate degradation I",
  "GLYCOLYSIS" = "Glycolysis I",
  "LACTOSECAT-PWY" = "Lactose and galactose degradation I",
  "NONMEVIPP-PWY" = "Methylerythritol phosphate (MEP) pathway I",
  "ORNARGDEG-PWY" = "Ornithine degradation I",
  "P108-PWY" = "Pyruvate fermentation to acetate",
  "P124-PWY" = "Bifidobacterium shunt (fructose-6-phosphate shunt)",
  "PANTO-PWY" = "Phosphopantothenate biosynthesis I",
  "PANTOSYN-PWY" = "Pantothenate biosynthesis I",
  "PEPTIDOGLYCANSYN-PWY" = "Peptidoglycan biosynthesis I",
  "PPGPPMET-PWY" = "ppGpp biosynthesis",
  "PRPP-PWY" = "PRPP biosynthesis I",
  "SER-GLYSYN-PWY" = "Superpathway of L-serine and glycine biosynthesis I",
  "TRNA-CHARGING-PWY" = "tRNA charging",
  "TRPSYN-PWY" = "L-tryptophan biosynthesis",
  "PWY0-166" = "Superpathway of peptidoglycan biosynthesis",
  "PWY0-1319" = "CDP-diacylglycerol biosynthesis II",
  "PWY0-1586" = "Peptidoglycan maturation (meso-diaminopimelate-containing)",
  "PWY-621" = "Sucrose degradation III",
  "PWY-801" = "Cobalamin salvage",
  "PWY-841" = "Heme biosynthesis II",
  "PWY-922" = "Mevalonate pathway I",
  "PWY-1042" = "Glycolysis IV",
  "PWY-2941" = "L-lysine biosynthesis I",
  "PWY-3841" = "Folate transformations II",
  "PWY-5097" = "Purine nucleotide de novo biosynthesis II",
  "PWY-5100" = "Pyruvate fermentation to acetate and lactate II",
  "PWY-5121" = "Geranylgeranyl diphosphate biosynthesis I",
  "PWY-5130" = "UMP biosynthesis",
  "PWY-5136" = "Fatty acid β-oxidation II (peroxisome)",
  "PWY-5188" = "Tetrapyrrole biosynthesis I (from glutamate)",
  "PWY-5367" = "Petroselinate biosynthesis",
  "PWY-5384" = "Sucrose degradation IV",
  "PWY-5392" = "reductive tricarboxylic acid (TCA) cycle II",
  "PWY-5484" = "Glycolysis II (from fructose 6-phosphate)",
  "PWY-5667" = "CDP-diacylglycerol biosynthesis I",
  "PWY-5686" = "UMP biosynthesis",
  "PWY-5910" = "Superpathway of heme biosynthesis",
  "PWY-5941" = "Glycogen degradation I",
  "PWY-6121" = "5-aminoimidazole ribonucleotide biosynthesis",
  "PWY-6122" = "5-aminoimidazole ribonucleotide biosynthesis II",
  "PWY-6123" = "IMP biosynthesis III",
  "PWY-6124" = "Adenosine nucleotide de novo biosynthesis II",
  "PWY-6125" = "Guanosine nucleotide de novo biosynthesis I",
  "PWY-6126" = "Adenosine nucleotide salvage I",
  "PWY-6143" = "UDP-N-acetylmuramoyl-pentapeptide biosynthesis I",
  "PWY-6151" = "S-adenosyl-L-methionine cycle",
  "PWY-6277" = "Superpathway of 5-aminoimidazole ribonucleotide biosynthesis",
  "PWY-6293" = "Superpathway of L-ascorbate degradation",
  "PWY-6385" = "peptidoglycan biosynthesis III (mycobacteria)",
  "PWY-6386" = "UDP-N-acetylmuramoyl-pentapeptide biosynthesis II",
  "PWY-6387" = "UDP-N-acetylmuramoyl-pentapeptide biosynthesis III",
  "PWY-6470" = "Peptidoglycan biosynthesis III",
  "PWY-6471" = "Peptidoglycan biosynthesis IV (Enterococcus faecium)",
  "PWY-6507" = "4-deoxy-L-threo-hex-4-enopyranuronate degradation",
  "PWY-6527" = "Starch biosynthesis",
  "PWY-6595" = "Superpathway of guanosine nucleotide de novo biosynthesis II",
  "PWY-6607" = "Guanosine nucleotide salvage",
  "PWY-6608" = "Guanosine nucleotide degradation III",
  "PWY-6609" = "Guanosine nucleotide degradation IV",
  "PWY-6612" = "Adenosine nucleotide salvage II",
  "PWY-6700" = "Queuosine biosynthesis",
  "PWY-6703" = "PreQ0 biosynthesis",
  "PWY-6823" = "Molybdenum cofactor biosynthesis",
  "PWY-6859" = "TCA cycle VIII (aerobic)",
  "PWY-6876" = "Isopropanol biosynthesis",
  "PWY-6936" = "Seleno-amino acid biosynthesis",
  "PWY-7031" = "Superpathway of purine nucleotide salvage",
  "PWY-7184" = "Pyrimidine deoxyribonucleotide de novo biosynthesis",
  "PWY-7187" = "Pyrimidine ribonucleotide de novo biosynthesis",
  "PWY-7197" = "Pyrimidine deoxyribonucleotide biosynthesis",
  "PWY-7198" = "Pyrimidine ribonucleotide salvage",
  "PWY-7199" = "Pyrimidine deoxyribonucleotide salvage",
  "PWY-7210" = "Pyrimidine deoxyribonucleotides biosynthesis from CTP",
  "PWY-7221" = "Guanosine nucleotide salvage II",
  "PWY-7228" = "Superpathway of adenosine nucleotide biosynthesis",
  "PWY-7229" = "Adenosine nucleotide biosynthesis",
  "PWY-7234" = "Inosine-5'-phosphate biosynthesis II",
  "PWY-7237" = "CDP-diacylglycerol biosynthesis III",
  "PWY-7238" = "Sucrose biosynthesis II",
  "PWY-7357" = "Thiamine diphosphate biosynthesis",
  "PWY-7391" = "Anaerobic energy metabolism",
  "PWY-7392" = "Aerobic respiration II",
  "PWY-7790" = "Pyridoxal 5'-phosphate biosynthesis I",
  "PWY-7791" = "Pyridoxal 5'-phosphate salvage",
  "PWY-7858" = "(5Z)-dodecenoate biosynthesis II",
  "PWY-7953" = "Glycogen biosynthesis I",
  "PWY-7977" = "Lipid IVA biosynthesis",
  "PWY-8178" = "UDP-N-acetylmuramoyl-pentapeptide biosynthesis IV",
  "PWY66-389" = "Purine nucleobase degradation",
  "PWY66-429" = "Allantoin degradation to glyoxylate III",  
  "P221-PWY" = "Octane oxidation",
  "COA-PWY-1" = "Coenzyme A biosynthesis II (mammalian)",
  "PWY-6629" = "Phosphatidylglycerol biosynthesis II",
  "GLYCOGENSYNTH-PWY" = "Glycogen biosynthesis I (from ADP-D-glucose)",
  "HSERMETANA-PWY" = "L-methionine biosynthesis via transsulfuration",
  "PWY-7851" = "Superpathway of coenzyme A biosynthesis",
  "PWY-6317" = "Galactose degradation I (Leloir pathway)",
  "HISTSYN-PWY" = "L-histidine biosynthesis",
  "PWY-6163" = "Chorismate biosynthesis from 3-dehydroquinate",
  "PWY-7208" = "Superpathway of pyrimidine nucleobase salvage",
  "PWY-5695" = "Purine nucleotide degradation",
  "PWY-7220" = "Adenosine deoxyribonucleotide de novo biosynthesis II",
  "PWY-7222" = "Guanosine deoxyribonucleotide de novo biosynthesis II",
  "1CMET2-PWY" = "N10-formyl-tetrahydrofolate biosynthesis",
  "PWY-7560" = "MEP Pathway II",
  "DTDPRHAMSYN-PWY" = "dTDP-L-rhamnose biosynthesis",
  "COLANSYN-PWY" = "Colanic acid biosynthesis", 
  "PROPFERM-PWY" = "Propionate fermentation",
  "PWY-8188" = "Stickland fermentation (glycine branch)",
  "PWY-8189" = "Stickland fermentation (proline branch)",
  "PWY-5494" = "Pyruvate fermentation to short-chain fatty acids",
  "PWY-6478" = "GDP-D-glycero-alpha-D-manno-heptose biosynthesis",
  "P41-PWY" = "Pyruvate fermentation to acetate and (S)-lactate I",
  "PWY-6138" = "CMP-N-acetylneuraminate biosynthesis I (eukaryotes)",
  "PWY-5345" = "L-methionine biosynthesis III",
  "PYRIDNUCSYN-PWY" = "NAD biosynthesis I (from aspartate)",
  "PWY-7316" = "dTDP-N-acetylviosamine biosynthesis",
  "FASYN-INITIAL-PWY" = "Fatty acid biosynthesis – initial steps",
  "RIBOSYN2-PWY" = "Riboflavin biosynthesis",
  "PWY-6897" = "Thiamine salvage II",
  "RHAMCAT-PWY" = "L-rhamnose degradation I",
  "PWY-6284" = "Superpathway of unsaturated fatty acids biosynthesis (E.coli)",
  "PWY-5659" = "GDP-mannose biosynthesis",
  "P562-PWY" = "Myo-inositol degradation I",
  "SULFATE-CYS-PWY" = "L-cysteine biosynthesis from sulfate",
  "PWY-1269" = "CMP-3-deoxy-D-manno-octulosonate biosynthesis I",
  "PWY-1861" = "Formaldehyde assimilation",
  "PWY-7688" = "dTDP-D-ravidosamine and dTDP-4-acetyl-D-ravidosamine biosynthesis",
  "COMPLETE-ARO-PWY" = "Superpathway of aromatic amino acid biosynthesis",
  "PWY1ZNC-1" = "Assimilatory sulfate reduction IV,",
  "PWY-5855" = "Ubiquinol-7 biosynthesis (prokaryotic)",
  "SO4ASSIM-PWY" = "Sulfate reduction I (assimilatory)",
  "PWY-7323" = "superpathway of GDP-mannose-derived O-antigen building blocks biosynthesis",
  "PWY-6285" = "superpathway of fatty acid biosynthesis",
  "BRANCHED-CHAIN-AA-SYN-PWY" = "Superpathway of branched-chain amino acid biosynthesis",
  "P185-PWY" = "Formaldehyde assimilation I",
  "ALLANTOINDEG-PWY" = "Allantoin degradation to glyoxylate",
  "PWY-7527" = "L-methionine salvage cycle III")

```

#1. Species specific differences for 16S, shotgun, SCFA, and metabolomics 
```{r}
#two way tests for taxonomic and functional div data
wilcox.test(otu_shannon ~ Species, data = shotgun_metadata)
effectsize::rank_biserial(otu_shannon ~ Species, data = shotgun_metadata)
shotgun_metadata %>% group_by(Species) %>% summarise(median(otu_shannon)) 

wilcox.test(otu_richness ~ Species, data = shotgun_metadata)
effectsize::rank_biserial(otu_richness ~ Species, data = shotgun_metadata)
shotgun_metadata %>% group_by(Species) %>% summarise(median(otu_richness)) 

wilcox.test(shannon ~ Species, data = shotgun_metadata)
effectsize::rank_biserial(shannon ~ Species, data = shotgun_metadata)

wilcox.test(observed_pathways ~ Species, data = shotgun_metadata)
effectsize::rank_biserial(observed_pathways ~ Species, data = shotgun_metadata)

wilcox.test(acetate/total_SCFA ~ Species, data = shotgun_metadata)
effectsize::rank_biserial(acetate/total_SCFA ~ Species, data = shotgun_metadata)

wilcox.test(valerate/total_SCFA ~ Species, data = shotgun_metadata)
effectsize::rank_biserial(valerate/total_SCFA ~ Species, data = shotgun_metadata)

wilcox.test(propionate/total_SCFA ~ Species, data = shotgun_metadata)
effectsize::rank_biserial(propionate/total_SCFA ~ Species, data = shotgun_metadata)

wilcox.test(butyrate/total_SCFA ~ Species, data = shotgun_metadata)
effectsize::rank_biserial(butyrate/total_SCFA ~ Species, data = shotgun_metadata)

#adding the 16S figures - shannon, richness for ASVs
fig2a.1 <- shotgun_metadata %>% ggplot( aes(x = Species, y = otu_shannon, color = Species)) +
  geom_boxplot() + geom_jitter(alpha = 0.3, size = 2) + xlab("") + ylab("Shannon diversity of ASVs") +
  theme_pubr(base_size = 24) + scale_color_bmj(name = NULL) + geom_signif(comparisons = list(c("SB", "SC")), map_signif_level = TRUE, color = "black",  test = "wilcox.test", step_increase = 0.1 ) +
  labs(title = " ") + theme(axis.text.x = element_blank()) 

fig2a.2 <- shotgun_metadata %>% ggplot( aes(x = Species, y = otu_richness, color = Species)) +
  geom_boxplot() + geom_jitter(alpha = 0.3, size = 2) + xlab("") + ylab("Number of observed ASVs") +
   scale_color_bmj() + theme_pubr(base_size = 24) + geom_signif(comparisons = list(c("SB", "SC")), map_signif_level = TRUE, color = "black",  test = "wilcox.test", step_increase = 0.1 ) +  labs(title = " ") + theme(legend.position = "none",, axis.text.x = element_blank()) 

fig2a.3 <- shotgun_metadata %>% ggplot( aes(x = otu_PC1, y = otu_PC2, color = Species)) +
  geom_point(alpha = 0.6, size = 4) + labs(title = " ") + xlab("PCoA1") + ylab("PCoA2") +
  theme_pubr(base_size = 24) + scale_color_bmj() 

fig2_block2a <- fig2a.1 + fig2a.2 + fig2a.3

#PERMANOVAs for 16S
#Aitchison PCA = PCA on clr-transformed relative abundances 
pseudocount            <- 0.65 #for clr-transformation
#alpha div measures
ftbl                   <- t(as(otu_table(ps),"matrix"))
feature_clr_abund      <- compositions::clr(t(as(otu_table(ps), "matrix") + pseudocount))
feature_pca_abund      <- prcomp(feature_clr_abund, center = TRUE, scale. = FALSE) # PCA on clr transformed data 

asv_metadata           <- readRDS("asv_metadata_Jun92026_newloc_type.rds") #update this too
aitchison_dist         <- dist(feature_clr_abund, method = "euclidean")
head(aitchison_dist)
nrow(aitchison_dist) 
nrow(asv_metadata)
table(asv_metadata$housing_type)

asv_metadata$housing_type     <- greg_data$new_location_type[match(as.numeric(asv_metadata$simple.id), greg_data$Animal)]
asv_metadata$housing_type[82] <-  "Naïve social group"
set.seed(123)
perm_asv <- adonis2(aitchison_dist ~ Species + Sex + Age + housing_type + enclosure_size,  data = asv_metadata, permutations = 9999, by = "margin")
summary(perm_asv) 

#dispersion tests
permutest(betadisper(aitchison_dist, asv_metadata$Species)) #significant
#As age is continuous, extract the distances from each sample to the overall group centroid and correlate those distances directly with your continuous Age variable
sample_distances <- betadisper(aitchison_dist, asv_metadata$Species)$distances
cor.test(sample_distances, asv_metadata$Age, method = "spearman") #non significant 

permutest(betadisper(aitchison_dist, asv_metadata$Sex)) #non significant 
permutest(betadisper(aitchison_dist, asv_metadata$housing_type)) #significant 
permutest(betadisper(aitchison_dist, asv_metadata$enclosure_size)) #non significant 

#asv results
adonis_asv          <- as.data.frame(perm_asv)
adonis_asv$Variable <- rownames(adonis_asv)
adonis_asv <- adonis_asv %>%
                        dplyr::filter(Variable %!in% c("Residual", "Total")) %>%
                        dplyr::select(Variable, R2,`Pr(>F)`) %>%
                        dplyr::mutate(Omic = "16S") %>% 
                        mutate(sig = case_when(`Pr(>F)` < 0.001 ~ "***",`Pr(>F)` < 0.01  ~ "**",`Pr(>F)` < 0.05  ~ "*", TRUE ~ ""))

adonis_asv 
adonis_asv$Variable[which(adonis_asv$Variable == "housing_type")]       <- c("Housing type")
adonis_asv$Variable[which(adonis_asv$Variable == "enclosure_size")]     <- c("Enclosure size")
adonis_asv$Variable <- factor(adonis_asv$Variable, levels = c("Species", "Age", "Sex", "Housing type", "Enclosure size"))

#create the plot
fig2_block2b <- ggplot(adonis_asv , aes(x = Variable, y = R2 * 100, fill = Variable)) +
  geom_col() + scale_fill_npg(alpha = 0.8) + 
  geom_text(aes(label = sig), hjust = -0.5, vjust = 0.2) +
  coord_flip() + facet_grid(.~ Omic) + ylim(c(0,6)) +
  theme_pubr(base_size = 24) + xlab("") +
  theme(legend.position = "none", strip.text = element_text(face = "bold")) +
  labs(x = NULL, y = expression("% Partial " * R^2 * "explained"), tag = "C", size = 24, face = "bold")

BMC_Fig2 <- plot_grid(fig2_block2a, fig2_block2b)
#ggsave( "BMC_BioFig2.jpeg", BMC_Fig2, width = 27, height = 8, dpi = 500, units = "in")

###Next adding the shotgun div figures for functional data
fig2b.1 <- shotgun_metadata %>% ggplot( aes(x = Species, y = shannon, color = Species)) +
  geom_boxplot() + geom_jitter(alpha = 0.3, size = 2) + xlab("") + ylab("Shannon diversity \nfunctional pathways") +
  theme_pubr(base_size = 24) + scale_color_bmj() +
  labs(title = " ") + theme(axis.text.x = element_blank())

fig2b.2 <- shotgun_metadata %>% ggplot( aes(x = Species, y = observed_pathways_new, color = Species)) +
  geom_boxplot() + geom_jitter(alpha = 0.3, size = 2) + xlab("") + ylab("Number of observed \nfunctional pathways") +
  theme_pubr(base_size = 24) + scale_color_bmj()  +
  labs(title = " ") + theme(axis.text.x = element_blank())
 
#Plot PCA
fig2b.3 <- ggplot(shotgun_metadata, aes(x = shotgun_atc_PCoA1, y = shotgun_atc_PCoA2, color = Species)) +
  geom_point(size = 4, alpha = 0.6) + scale_color_bmj()  +
  labs(title = " ")  + xlab("PCoA1") + ylab("PCoA2") +
  theme_pubr(base_size = 24)  

Fig3_block1 <- plot_grid(fig2b.1, fig2b.2, fig2b.3, ncol = 3, nrow = 1)

#Permanova results 
pathway_aitch_dist <- readRDS("sbol_pathway_aitchison.rds")
head(pathway_aitch_dist)

perm_func         <- adonis2(pathway_aitch_dist ~ Species + Age + Sex + `Housing type` + `Enclosure size`,
  data = shotgun_metadata,
  permutations = 9999,
  by = "margin")
perm_func

#dispersion tests
permutest(betadisper(pathway_aitch_dist, shotgun_metadata$Species))
permutest(betadisper(pathway_aitch_dist, shotgun_metadata$Age))
permutest(betadisper(pathway_aitch_dist, shotgun_metadata$`Housing type`)) #all ok here
permutest(betadisper(pathway_aitch_dist, shotgun_metadata$Sex)) 
permutest(betadisper(pathway_aitch_dist, shotgun_metadata$`Enclosure size`)) 


#function results
adonis_func          <- as.data.frame(perm_func)
  adonis_func$Variable <- rownames(adonis_func)
  adonis_func <- adonis_func %>%
    dplyr::filter(Variable %!in% c("Residual", "Total")) %>%
    dplyr::select(Variable, R2,`Pr(>F)`) %>%
    dplyr::mutate(Omic = "functional pathways") %>% mutate(sig = case_when(`Pr(>F)` < 0.001 ~ "***",`Pr(>F)` < 0.01  ~ "**",`Pr(>F)` < 0.05  ~ "*", TRUE ~ ""))
  
adonis_func$Variable[which(adonis_func$Variable ==  "`Housing type`")]     <- "Housing type"
adonis_func$Variable[which(adonis_func$Variable ==  "`Enclosure size`")]   <- "Enclosure size"
 
  
adonis_func$Variable <- factor(adonis_func$Variable, levels = c("Species", "Age", "Sex", "Housing type", "Enclosure size"))
Fig3_block2 <- ggplot(adonis_func , aes(x = (Variable), y = R2 * 100, fill = Variable)) +
  geom_col() + scale_fill_npg(alpha = 0.8) +
  geom_text(aes(label = sig), hjust = -0.5, vjust = 0.2) +
  coord_flip() + facet_grid(.~ Omic) + ylim(c(0,6)) +
  theme_pubr(base_size = 24) + 
  theme( legend.position = "none", strip.text = element_text(face = "bold")) +
  labs(x = NULL,y = expression("% Partial " * R^2 * "explained"))  

#add SCFA next
fig3c.1 <- shotgun_metadata %>% ggplot(aes(x = Species, y = acetate/total_SCFA, color = Species)) + geom_violin()  + geom_jitter(size = 4, alpha = 0.6) +
  theme_pubr(base_size = 24) + scale_color_bmj()  + xlab("") + ylab("prop of acetate") + theme(legend.position = "none") 

fig3c.2 <- shotgun_metadata %>% ggplot(aes(x = Species, y = butyrate/total_SCFA, color = Species)) + geom_violin()  + geom_jitter(size = 4, alpha = 0.6) +
  theme_pubr(base_size = 24) + scale_color_bmj()  + xlab("") + ylab("prop of butyrate") + theme(legend.position = "none") 

fig3c.3 <- shotgun_metadata %>% ggplot(aes(x = Species, y = propionate/total_SCFA, color = Species)) + geom_violin()  + geom_jitter(size = 4, alpha = 0.6) +
  theme_pubr(base_size = 24) + scale_color_bmj()  + xlab("") + ylab("prop of propionate") + theme(legend.position = "none") 

fig3c.4 <- shotgun_metadata %>% ggplot(aes(x = Species, y = valerate/total_SCFA, color = Species)) + geom_violin()  + geom_jitter(size = 4, alpha = 0.6) +
  theme_pubr(base_size = 24) + scale_color_bmj()  + xlab("") + ylab("prop of isovalerate") + theme(legend.position = "none")  

Fig3_block3 <- (fig2c.1 + fig2c.2) / (fig2c.3 + fig2c.4)
Fig3_block3

#add metabolites next - PCA and permanova
metadata      <- read.csv("metadata_for_metabolomics.csv", row.names = 1, check.names = FALSE)
metab_pareto  <- readRDS("sbol_pareto_scaled_metabolites.rds")
metab_nmds    <- metab_pareto
#Run NMDS
set.seed(123)
nmds_res <- metaMDS(metab_nmds, distance = "bray",   # or "euclidean"
  k = 2,
  trymax = 100,
  autotransform = FALSE)
#Stress value
nmds_res$stress

nmds_df                 <- as.data.frame(scores(nmds_res))
nmds_df$normalized_name <- rownames(nmds_df)
nmds_df                 <- left_join(nmds_df, metadata, by = "normalized_name")

#plot
Fig3_block4 <- nmds_df %>% filter(!is.na(Species)) %>%  ggplot(aes(NMDS1, NMDS2, color = Species)) +
  geom_point(size = 3, alpha = 0.6) +
  theme_pubr(base_size = 24) + stat_ellipse(level = 0.95,
               geom = "polygon",
               alpha = 0.02) + labs(title = "Untargeted metabolomics") + scale_color_bmj()
Fig3_block4

perm_metab <- adonis2(metab_dist ~ Species + Age + Sex + `Housing type` + `Enclosure size`, data = shotgun_metadata, permutations = 9999, by = "margin")
perm_metab

#function results
adonis_metab          <- as.data.frame(perm_metab)
  adonis_metab$Variable <- rownames(adonis_metab)
  adonis_metab <- adonis_metab %>%
    dplyr::filter(Variable %!in% c("Residual", "Total")) %>%
    dplyr::select(Variable, R2,`Pr(>F)`) %>%
    dplyr::mutate(Omic = "metabolites")  %>% mutate(sig = case_when(`Pr(>F)` < 0.001 ~ "***",`Pr(>F)` < 0.01  ~ "**",`Pr(>F)` < 0.05  ~ "*", TRUE ~ ""))

adonis_metab$Variable[which(adonis_metab$Variable ==  "`Housing type`")]     <- "Housing type"
adonis_metab$Variable[which(adonis_metab$Variable ==  "`Enclosure size`")]   <- "Enclosure size"
 
  
adonis_metab$Variable <- factor(adonis_metab$Variable, levels = c("Species", "Age", "Sex", "Housing type", "Enclosure size"))

Fig3_block5 <- ggplot( adonis_metab , aes(x = (Variable), y = R2 * 100, fill = Variable)) +
  geom_col()  + scale_fill_npg(alpha = 0.8) +
  geom_text(aes(label = sig), hjust = -0.5, vjust = 0.2) +
  coord_flip() + facet_grid(.~ Omic) +
  theme_pubr(base_size = 24) + ylim(c(0,6)) +
  theme( legend.position = "none", strip.text = element_text(face = "bold")) +
  labs(x = NULL,y = expression("% Partial " * R^2 * "explained"))  

#dispersion tests
permutest(betadisper(metab_dist, shotgun_metadata$Species))
permutest(betadisper(metab_dist, shotgun_metadata$Age))
permutest(betadisper(metab_dist, shotgun_metadata$Location_Type)) #only location is significant
permutest(betadisper(metab_dist, shotgun_metadata$Sex)) 
permutest(betadisper(metab_dist, shotgun_metadata$enclosure)) 


Fig3 <- (Fig3_block1 | Fig3_block2) / (Fig3_block3 | Fig3_block4 | Fig3_block5) + plot_layout(guides = "collect")
#ggsave("BMC_BioFig3.jpeg", Fig3, dpi = 300, width = 28, height = 14, units = "in")
```

#2.1. Differential abundance of taxa
```{r}
#Testing differential abundances (16S and shotgun data) using Maaslin3 - prep the maaslin3 metadata for both shotgun and 16S and then just feed into this chunk
genus_rel            <- read.csv("genus_abundance_for_MaAsLin3.csv")
metadata_maaslin_asv <- readRDS("metadata_maaslin_asv.RDS")
#save this as an RDS object 
saveRDS(metadata_maaslin_asv, "metadata_maaslin_asv.RDS")
metadata_maaslin_asv[metadata_maaslin_asv$Location_type != "Individual",]

table(metadata_maaslin_asv$sample.id)
metadata_maaslin_asv <- sample_data(ps_rel)
metadata_maaslin_asv <- data.frame(metadata_maaslin_asv)
head(shotgun_metadata) #add the right age group.
metadata_maaslin_asv$Age_group        <- shotgun_metadata$Age_group1[match(metadata_maaslin_asv$sample.id, asv_metadata$sample.id)] 
metadata_maaslin_asv$Age_group        <- factor(metadata_maaslin_asv$Age_group, levels = c("Infant", "Yearling", "Juvenile", "Adult", "Geriatric", "Geriatric_old"))
metadata_maaslin_asv$Location_Type    <- asv_metadata$Location_type[match(metadata_maaslin_asv$sample.id, asv_metadata$sample.id)]
metadata_maaslin_asv$enclosure        <- asv_metadata$enclosure[match(metadata_maaslin_asv$sample.id, asv_metadata$sample.id)]

#Clean metadata
metadata_maaslin_asv$Species           <- factor(metadata_maaslin_asv$Species)
metadata_maaslin_asv$Sex               <- factor(metadata_maaslin_asv$Sex)
metadata_maaslin_asv$Location_Type     <- factor(metadata_maaslin_asv$Location_Type, levels = c("Naïve social group","Experimental group","Treatment group","Individual"))

metadata_maaslin_asv$Location_Type     <- relevel(metadata_maaslin_asv$Location_Type, ref = "Experimental group")
metadata_maaslin_asv$enclosure         <- factor(metadata_maaslin_asv$enclosure)
metadata_maaslin_asv$enclosure         <- relevel(metadata_maaslin_asv$enclosure, ref = "Small")
metadata_maaslin_asv                   <- metadata_maaslin_asv[which(!is.na(metadata_maaslin_asv$Species)),]
metadata_maaslin_asv$Age               <- asv_metadata$Age[match(metadata_maaslin_asv$sample.id, asv_metadata$sample.id)] 

str(metadata_maaslin_asv)

#metadata_maaslin_asv$Age_group[82] <- "Infant"
metadata_maaslin_asv$Age[82] <- 0

metadata_maaslin_asv[82,]
metadata_maaslin_asv$Age_group[which(is.na(metadata_maaslin_asv$Age_group))] <- "Geriatric"
metadata_maaslin_asv$Age[which(is.na(metadata_maaslin_asv$Age))] <- 20
rownames(metadata_maaslin_asv)
rownames(filtered_genus_table)
rownames(genus_rel)
head(metadata_maaslin_asv)
library(maaslin3)
#remove the unknown taxa from genus_rel file

genus_rel <- genus_rel %>% dplyr::select(-Unknown) %>% as.data.frame() #removing teh unknown genus
#Species and age only
fit_species_age_group_both_species <- maaslin3(input_data = genus_rel,
                                          input_metadata = metadata_maaslin_asv,
                                          output = "maaslin3_species_age_genus_abun_Aug7",
                                          formula  = "~ Species + Age", 
                                          normalization = "TSS", 
                                          transform = "LOG",
                                          standardize = FALSE, 
                                          max_significance = 0.1,
                                          max_pngs = 100,  evaluate_only = "abundance", warn_prevalence = F,
                                          cores = 1, save_models = TRUE)


results                               <- read.delim("maaslin3_species_age_genus_abun_Aug7/all_results.tsv") #3 genus only with abundance only selection
abundance_results                     <- subset(results, model == "abundance")

abundance_results %>% filter(metadata == "Species", qval_joint < 0.05) %>% as.data.frame() %>% nrow()
abundance_results %>% filter(metadata == "Age", qval_joint < 0.05) %>% as.data.frame() %>% nrow()

abundance_results$direction           <- ifelse(abundance_results$coef < 0, "lower in SC", "higher in SC")
abundance_results$log_qval_individual <- -log10(abundance_results$qval_individual)
-log10(0.05)
species_results             <- abundance_results %>% filter(metadata == "Species", qval_joint < 0.05) %>% as.data.frame()
species_results$direction   <- ifelse(species_results$coef < 0, "lower in SC", "higher in SC")
species_results$direction   <- factor(species_results$direction, levels = c("lower in SC", "higher in SC"))

#keep significant hits only
sig_results <- species_results %>%
  filter(qval_joint < 0.05) 

#Reorder taxa/genera by coefficient size
sig_results <- sig_results %>%
  mutate(feature = fct_reorder(feature, coef), direction = ifelse(coef < 0, "lower in SC", "higher in SC"))

#Plot only sig
genus_volcano_plot <- abundance_results %>% filter(!is.na(qval_individual) & metadata  == "Species") %>% ggplot(aes(x = coef, y = log_qval_individual)) +
  geom_point(aes(color = direction),
             size = 4, alpha = 0.6) +
  geom_vline(xintercept = 0,
             linetype = "dashed") +  scale_fill_manual(values = c("lightblue","coral"),
    name = "qval_joint < 0.05") + geom_hline(yintercept = 1.3, linetype = "dotted", color = "maroon") +
  theme_pubr(base_size = 24) +
  labs( x = "Beta coefficient",
    y = "-log10(q-value)",
    title = "ASV abundance associations (by host species)") + theme(plot.title = element_text(hjust = 0.5, size = 20)) 
genus_volcano_plot #Fig 4A


genus_diff_maaslin <- ggplot(sig_results, aes(x = coef, y = feature, fill = direction)) +
  geom_col() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("lightblue", "coral"),
    name = "direction of effect") +
  labs(title = "Differential abundance associations",
    x = "Model coefficient (beta)",
    y = "Genus") +  theme_pubr(base_size = 24)  + theme(panel.grid.major.y = element_blank(), legend.position = "right")

genus_diff_maaslin #Fig 4B

abundance_results$direction <- factor(abundance_results$direction, levels = c("lower in SC", "higher in SC"))

write.csv(sig_results, "genus_maaslin_species.csv" , row.names = F)

#Species and all other covariates 
fit_species_allcovariates         <- maaslin3(input_data = genus_rel,
                                          input_metadata = metadata_maaslin_asv[which(metadata_maaslin_asv$Species == "SB"),] , #for boliviensis only metadata_maaslin_asv[which(metadata_maaslin_asv$Species == "SB"),] result is in "maaslin3_sbspecies_genus_asv_checkall_Aug6"
                                          output = "maaslin3_species_genus_asv_SBonlycheckall_Aug12",
                                          formula  = "~ Age + Sex ", 
                                          normalization = "TSS", 
                                          transform = "LOG",
                                          standardize = FALSE, 
                                          max_significance = 0.1,
                                          evaluate_only = "abundance", warn_prevalence = F,max_pngs = 100,
                                          cores = 1, save_models = TRUE)
#nothing is significant when including all covariates

```

#2.2. Differential abundance of functional pathways across the two species.
```{r}
#SHOTGUN MAASLIN3
filtered_pathway          <- readRDS("/Users/sharmisen/Desktop/POSTDOC PROJECTS/NORTHWESTERN/Squirrel Monkey Project/sbol_filtered_pathway_dataJul12026.rds")
filtered_pathway <- filtered_pathway[-87,] #removing the repeat 6524
head(filtered_pathway)
filtered_pathway_metadata <- shotgun_metadata
which(rownames(filtered_pathway) == "6524-SBOL-USA-22-LW-SH-1_S228_knead_merged_Abundance")

rownames(filtered_pathway) <- substr(rownames(filtered_pathway), 1,19)
rownames(filtered_pathway_metadata)

head(filtered_pathway_metadata)

str(filtered_pathway_metadata)
rownames(filtered_pathway_metadata) 
#changing categorical variables to factors and assinging references
filtered_pathway_metadata$Location_Type <- factor(filtered_pathway_metadata$Location_Type, levels = c("Naïve social group", "Experimental group", "Treatment group", "Individual" ))
filtered_pathway_metadata$Location_Type <- relevel(filtered_pathway_metadata$Location_Type, ref = "Naïve social group")
filtered_pathway_metadata$enclosure     <- factor(filtered_pathway_metadata$enclosure, levels = c("Small", "Medium", "Large", "Individual"))
filtered_pathway_metadata$enclosure     <- relevel(filtered_pathway_metadata$enclosure, ref = "Small")

#DO SHOTGUN MAASLIN FOR both SPECIES - to report species level differences only first
#All covariates
fit_pathway_all <- maaslin3(input_data = filtered_pathway,
                                          input_metadata = filtered_pathway_metadata,
                                          output = "maaslin_all_covariates_pathway_bothabunprevAug7",
                                          formula  = "~ Species + Age + Location_Type + enclosure ", 
                                          normalization = "TSS", 
                                          transform = "LOG",
                                          standardize = FALSE, 
                                          max_significance = 0.1,
                                          max_pngs = 100,
                                          cores = 1, 
                                          save_models = TRUE)

results_path          <- read.delim("maaslin_all_covariates_pathway_bothabunprevAug7/all_results.tsv") #just pull out species here first

abundance_results_func <- subset(results_path, model == "abundance")
abundance_results_func$log_qval_individual <- -log10(abundance_results_func$qval_individual)
abundance_results_func$color <- ifelse(abundance_results_func$coef < 0 , "lower in SC", "higher in SC")
abundance_results_func$color <- factor(abundance_results_func$color, level = c("lower in SC", "higher in SC"))
#Optional: keep significant hits only
sig_results_funct      <- abundance_results_func %>% filter(qval_joint < 0.05)
head(sig_results_funct)

sig_results_funct$PathwayID   <- str_remove(sig_results_funct$feature, ":.*")
sig_results_funct$PathwayName <- pathway_name_map[sig_results_funct$PathwayID]

sig_results_funct %>% filter(metadata == "Species" & !is.na(qval_individual) & qval_individual < 0.05  ) %>% nrow() #52
sig_results_funct %>% filter(metadata == "Location_Type" & !is.na(qval_individual) & qval_individual < 0.05 ) %>% nrow() #8
sig_results_funct %>% filter(metadata == "Age" & !is.na(qval_individual) & qval_individual < 0.05 ) %>% nrow() #79
sig_results_funct %>% filter(metadata == "enclosure" & !is.na(qval_individual) & qval_individual < 0.05 ) %>% nrow() #5
#subset by species
sig_results_funct_species  <- sig_results_funct %>% filter(metadata == "Species" & !is.na(qval_individual) & qval_individual < 0.05) %>% as.data.frame()
#Reorder by coefficient size
sig_results_funct_species$color    <- ifelse(sig_results_funct_species$coef < 0, "lower in SC", "higher in SC")
sig_results_funct_species          <- sig_results_funct_species %>% arrange(coef) %>% as.data.frame()
sig_results_funct_species$color    <- factor(sig_results_funct_species$color, level = c( "lower in SC", "higher in SC"))

#Plot only sig # do separate for  "species"

volcano_plot_sig_func_species <- abundance_results_func  %>% filter(metadata == "Species" & !is.na(coef)) %>% ggplot(aes(x = coef, y = log_qval_individual, col = color)) +
geom_jitter(size = 4, alpha = 0.75) + geom_hline(yintercept = 1.3, linetype = "dotted", color = "maroon") +
  geom_vline(xintercept = 0,
             linetype = "dashed") +  scale_fill_manual(values = c( "lightblue","coral"),  name = "qval_joint < 0.05") + 
  theme_pubr(base_size = 24) +
  labs( x = "Beta coefficient",
    y = "-log10(q-value)",
    title = "functional pathway abundance\n  associations in SC") + theme(plot.title = element_text(hjust = 0.5))
volcano_plot_sig_func_species #BMC_BioFig4C

sig_results_funct_species_fig <- sig_results_funct_species %>% filter(!is.na(qval_individual) & qval_individual < 0.05) %>% 
  arrange(coef) %>% slice_head(n = 30) %>%  ggplot(aes(x = coef, y = reorder(PathwayName, coef), fill = color)) + 
  geom_col() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("coral","lightblue"),
    name = "direction of effect") + 
  labs(title = "Differentially abundant functional pathways in SC",
    x = "Model coefficient (beta)",
    y = "Pathway") +  theme_pubr(base_size = 24)  + theme(panel.grid.major.y = element_blank(), legend.position = "right", plot.title = element_text(hjust = 0.5, size = 24))
sig_results_funct_species_fig

#ggsave( "BMC_BioFig4D", sig_results_funct_species_fig, width = 18, height = 12, units = "in", dpi = 440) 

sig_species_table <- sig_results_funct_species %>% as.data.frame()
write.csv(sig_species_table, "maaslin3_sig_func_species_table.csv", row.names = F)
 
abundance_results_func %>% filter(metadata == "Location_Type" & !is.na(qval_individual) & qval_individual < 0.05 & coef > 0) %>% nrow() #33
abundance_results_func %>% filter(metadata == "Location_Type" & !is.na(qval_individual) & qval_individual < 0.05 & coef < 0) %>% nrow() #4

sig_results_funct_loc$direction_of_effect <- ifelse(sig_results_funct_loc$coef > 0, "Higher than Breeding Group", "Lower than Breeding Group")
write.csv(sig_results_funct_loc, "sig_results_loc.csv", row.names = F)
```
#2.3. Differential abundance of metabolites across two species
```{r}
metabolite_matrix <- t(metab_pareto)
head(metabolite_matrix)
#add the metadata variables of interest 
metab_metadata <- read.csv("limma_metab_metadata_for_metabolites.csv")
head(metab_metadata)
metab_metadata$Species <- factor(metab_metadata$Species, levels = c("SC", "SB"))
metab_metadata$Species <- relevel(metab_metadata$Species, ref = "SB")
levels(metab_metadata$Species)
#add the updated enclosure and location
metabolite_matrix_check <- metabolite_matrix[, which(colnames(metabolite_matrix) %in% metab_metadata$SampleName) ]
ncol(metabolite_matrix_check) == nrow(metab_metadata)
#order metab_metadata according to the metab matrix 
metab_metadata <- metab_metadata[match(colnames(metabolite_matrix_check), metab_metadata$SampleName), ]
all(colnames(metabolite_matrix_check) == metab_metadata$normalized_name)

design <- model.matrix( ~ Species + Age  + Sex + Location_Type + enclosure,  data = metab_metadata)
# Now:
# rows = metabolites
# columns = samples

fit <- lmFit(metabolite_matrix_check, design)
fit <- eBayes(fit)
summary(fit)
species_res <- topTable(fit, coef = "SpeciesSC", number = Inf, adjust.method = "BH")

species_res$significant <- ifelse(species_res$adj.P.Val < 0.05, "significant", "ns")
species_res %>% filter(significant == "significant") %>% as.data.frame()

ggplot(species_res, aes(x = logFC,   y = -log10(adj.P.Val), color = significant)) + geom_hline(yintercept = 1.3, linetype = "dotted", color = "maroon") +
  geom_point(alpha = 0.75) + theme_pubr(base_size = 24) + scale_color_manual(values = c("lightblue","coral"),  name = "qval_joint < 0.05") +
  labs(x = "LogFC", y = "-log10(FDR)", title = "differential metabolite abundance associations \n in SC") +  
  theme(plot.title = element_text(hjust = 0.5))

sig_species <- rownames(species_res)[species_res$adj.P.Val < 0.05]
pheatmap(metabolite_matrix[sig, ], annotation_col = metab_metadata[, c("Species", "Sex", "Location_Type")]) #this is not working

age_res <- topTable(fit, coef = "Age", number = Inf, adjust.method = "BH")
head(age_res)
age_res$significant <- ifelse(age_res$adj.P.Val < 0.05, "significant", "ns")

ggplot(age_res, aes(x = logFC,   y = -log10(adj.P.Val), color = significant)) +
  geom_point(alpha = 0.75, size = 4) + 
  theme_pubr(base_size = 24) + scale_color_manual(values = c("coral", "lightblue"),  name = "pval_joint < 0.05") +
  geom_hline(yintercept = 1.3, linetype = "dotted", color = "maroon") + theme(legend.position = "right") +
  labs(x = "LogFC", y = "-log10(FDR)", title = "untargeted metabolite abundance \n associations (with age)") + theme(plot.title = element_text(hjust = 0.5)) 

#save the sig age related data
head(age_res)
report_file_metab         <- age_res %>% filter(significant == T)  %>% as.data.frame()
report_file_metab$feature <- rownames(report_file_metab)
write.csv(report_file_metab, "limma_metab_table.csv", row.names = F)
```

#5. Differential abundance of pathways related to GBMs across the two species 
```{r}
func_mat_gbm           <- readRDS("sbol_gbm_data_for_maaslin3_Aug10.rds")
maaslin_metadata       <- readRDS( "sbol_GBM_metadata_maaslin_Aug112926.RDS")
head(maaslin_metadata)
str(maaslin_metadata)
table(maaslin_metadata$Sample)

maaslin_output   <- maaslin3(input_data = func_mat_gbm,
                           input_metadata = maaslin_metadata,
                           output = 'maaslin_GBM_Aug10_all_covariates_abund_only', 
                           formula = '~ Species + Age + Location_Type + enclosure + log_reads', 
                           normalization = 'TSS',
                           transform = 'LOG',
                           max_significance = 0.1, evaluate_only = "abundance", warn_prevalence = F,
                           max_pngs = 100,
                           cores = 1,
                           save_models = TRUE) 

sig_gbms <- fread("maaslin_GBM_Aug10_all_covariates_abund_only/significant_results.tsv")

#species specific features
sp_feat <- sig_gbms$feature[sig_gbms$metadata == "Species"]
#do TSS for the func mat pathways
func_tss <- sweep(func_mat_gbm,1,rowSums(func_mat_gbm),"/")

#pathway of interest
pos                <- which(colnames(func_mat) == sp_feat[3]) #change number here 1-3 to get the plots
pathway_tss <- func_tss[, pos]

pseudo   <- min(pathway_tss[pathway_tss > 0], na.rm = TRUE) / 2

pathway_log2       <- log2(pathway_tss + pseudo)
values             <- pathway_log2

#create a plot for each of these features now
sampleid           <- rownames(func_mat)
temp_data          <- as.data.frame(cbind(sampleid, values))
temp_data$species  <- maaslin_metadata$Species[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$values   <- as.numeric(temp_data$values)
temp_data$age      <- maaslin_metadata$Age[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$sex      <- maaslin_metadata$Sex[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group  <- maaslin_metadata$Age_group1[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group2 <- maaslin_metadata$Age_Group[match(temp_data$sampleid, maaslin_metadata$Sample)]

str(temp_data)
head(temp_data)
which(is.na(temp_data$species))
hist(temp_data$values)
temp_data <- temp_data %>% filter(!is.na(species)) %>% as.data.frame()

#temp_data <- temp_data %>% dplyr::filter(values > 0) %>% as.data.frame()
gbm      <- lm(values ~ species + age, data = temp_data)
summary(gbm)

estradiol <- temp_data %>%  ggplot(aes(x = species, y = values,  color = species)) + geom_boxplot() + 
  geom_jitter(alpha = 0.6, size = 4) + xlab("") + theme_pubr(base_size = 24) + theme(legend.position = "right", axis.text.x = element_blank(),plot.title = element_text(hjust = 0.5)) +    
  scale_color_bmj() + ylab("pathway abundance \n  (TSS normalized:log2 transformed)") + xlab("")  +  ggtitle(sp_feat[1])  + xlab(" ") + 
  geom_signif(comparisons = list(c("SB", "SC")), map_signif_level = TRUE, color = "black",  test = "wilcox.test", step_increase = 0.1 )

glutamate <- temp_data %>%  ggplot(aes(x = species, y = values,  color = species)) + geom_boxplot() + 
  geom_jitter(alpha = 0.6, size = 4) + xlab("") + theme_pubr(base_size = 24) + theme(legend.position = "right", axis.text.x = element_blank(), plot.title = element_text(hjust = 0.5)) +    
  scale_color_bmj() + ylab(" ") + xlab("")  +  ggtitle(sp_feat[2])  + xlab(" ") + 
  geom_signif(comparisons = list(c("SB", "SC")), map_signif_level = TRUE, color = "black",  test = "wilcox.test", step_increase = 0.1 )

quinol <- temp_data %>%  ggplot(aes(x = species, y = values,  color = species)) + geom_boxplot() + 
  geom_jitter(alpha = 0.6, size = 4) + xlab("") + theme_pubr(base_size = 24) + theme(legend.position = "right", axis.text.x = element_blank(), plot.title = element_text(hjust = 0.5)) +    
  scale_color_bmj() + ylab(" ") + xlab("")  +  ggtitle(sp_feat[3])  + xlab(" ") + 
  geom_signif(comparisons = list(c("SB", "SC")), map_signif_level = TRUE, color = "black",  test = "wilcox.test", step_increase = 0.1 )

estradiol | quinol | glutamate #Fig 5 species GBM differences

```

#6. Age specific patterns in boliviensis only. Modeling taxonomic and functional shannon diversity GAMs. 
```{r}
boliv_color <- c("#2A6EBBFF")
str(shotgun_metadata)
#Age related patterns in the taxonomic composition and functional pathway diversity (shannon diversity)
shotgun_metadata$Species       <- factor(shotgun_metadata$Species)
shotgun_metadata$Sex           <- factor(shotgun_metadata$Sex)
shotgun_metadata$asv_log_reads <- log(shotgun_metadata$asv_reads)

m_linear <- gam(otu_shannon ~  Age + Sex + enclosure + Location_Type, data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "REML")
m_smooth <- gam(otu_shannon ~  s(Age) + Sex + enclosure + Location_Type, data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "REML")

anova(m_linear, m_smooth, test = "F")
summary(m_smooth)

AIC(m_linear, m_smooth)
hist(shotgun_metadata$otu_shannon)

#no non-linear effect of age 
g0 <- gam(otu_shannon ~ s(Age), data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")
g1 <- gam(otu_shannon ~ s(Age) +  Sex +  Location_Type + enclosure, data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")
g2 <- gam(otu_shannon ~ s(Age) +  Sex +  enclosure, data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")
g3 <- gam(otu_shannon ~ s(Age) +  Sex +  Location_Type, data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")
g4 <- gam(otu_shannon ~ s(Age) +  Sex , data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")

bbmle::AICctab(g0, g1, g2, g3,g4,  weights = T, base = T) #g0 and g2 was best model 
summary(g0)

final_asv_shannon <- gam(otu_shannon ~ s(Age),data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "REML")
summary(final_asv_shannon)

newdat            <- expand.grid(Age = seq(min(shotgun_metadata$Age[shotgun_metadata$Species == "SB"], na.rm=TRUE),
                                max(shotgun_metadata$Age[shotgun_metadata$Species == "SB"], na.rm=TRUE),length.out = 100))
pred              <- predict(g0, newdata = newdat, se.fit = TRUE)

newdat$fit        <- pred$fit
newdat$se         <- pred$se.fit
newdat$upper      <- newdat$fit + 1.96 * newdat$se
newdat$lower      <- newdat$fit - 1.96 * newdat$se

#plot smooth aging curves
fig3.a <- shotgun_metadata %>% filter(Species == "SB") %>% ggplot() + geom_point(aes(x = Age, y = otu_shannon, color= boliv_color) , alpha = 0.4) +
        geom_line(data = newdat, aes(x = Age, y = fit, color= boliv_color) , size = 2, col = "black") + 
        geom_ribbon(data = newdat, aes(x = Age, ymin = lower, ymax = upper), alpha = 0.3, color = "black", linewidth = 0.4) + 
        scale_fill_manual(values = boliv_color, name = NULL) +
        theme_pubr(base_size = 24) + scale_color_manual(values = boliv_color, name = NULL) +
        labs(x = "", y = "Shannon Diversity \n of ASVs") + guides(color = "none") 
fig3.a

#doing the same for richness now
m_rich_linear <- gam(otu_richness ~ Species + Age + Sex + enclosure + Location_Type, data = shotgun_metadata, method = "REML")
m_rich_smooth <- gam(otu_richness ~ Species + s(Age) + Sex + enclosure + Location_Type, data = shotgun_metadata, method = "REML")

AIC(m_rich_linear, m_rich_smooth) #no need for smooth term for age

#no non-linear effect of age 
r0 <- gam(otu_richness ~ s(Age), data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")
r1 <- gam(otu_richness ~ s(Age) +  Sex +  Location_Type + enclosure, data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")
r2 <- gam(otu_richness ~ s(Age) +  Sex +  enclosure, data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")
r3 <- gam(otu_richness ~ s(Age) +  Sex +  Location_Type, data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")

bbmle::AICctab(r0, r1, r2, r3, weights = T, base = T) #r0 and r2 was best model 
summary(r0)

final_asv_richness <- gam(otu_richness ~ s(Age),
          data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "REML")
summary(final_asv_richness)

newdat2 <- expand.grid(Age = seq(min(shotgun_metadata$Age[shotgun_metadata$Species == "SB"], na.rm=TRUE),
            max(shotgun_metadata$Age[shotgun_metadata$Species == "SB"], na.rm=TRUE), length.out=100))

pred2 <- predict(r0, newdata = newdat2, se.fit = TRUE)

newdat2$fit <- pred2$fit
newdat2$se  <- pred2$se.fit
newdat2$upper <- newdat2$fit + 1.96 * newdat2$se
newdat2$lower <- newdat2$fit - 1.96 * newdat2$se

#plot smooth aging curves
fig3.b <- ggplot() + geom_point(data = shotgun_metadata[shotgun_metadata$Species == "SB",], aes(x = Age, y = otu_richness),
             alpha = 0.4, color = boliv_color) +
  geom_line(data = newdat2, aes(x = Age, y = fit, color = boliv_color) , size = 2, col = "black") + 
  geom_ribbon(data = newdat2, aes(x = Age, ymin = lower, ymax = upper),alpha = 0.3, color = "black", linewidth = 0.4) + scale_fill_manual(values = boliv_color, name = NULL) +
  theme_pubr(base_size = 24) + scale_color_manual(values = boliv_color, name = NULL) +
  labs(x = "Age (years)", y = "Observed ASVs", ) + labs(color = NULL, ) + ylim(c(100,500)) + guides(color = "none")

fig3 <- fig3.a / fig3.b
fig3

#GAMs for functional data Naive social groups
#shannon diversity
m_linear <- gam(shannon ~  Age + Sex + enclosure + Location_Type, data = shotgun_metadata[shotgun_metadata$Species == "SB",],method = "REML")
m_smooth <- gam(shannon ~  s(Age) + Sex + enclosure + Location_Type, data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "REML")

anova(m_linear, m_smooth, test = "F")

AIC(m_linear, m_smooth)#, Species, bs = "fs"

sh0 <- gam(shannon ~ s(Age) , data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")
sh1 <- gam(shannon ~ s(Age) + Sex  + Location_Type, data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")
sh2 <- gam(shannon ~ s(Age) + Sex + enclosure + Location_Type  , data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")
sh3 <- gam(shannon ~ s(Age) + Sex + enclosure , data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")

bbmle::AICctab(sh0, sh1, sh2, sh3, weights = T, base = T)
summary(sh1)
newdat3 <- expand.grid( Age = seq(min(shotgun_metadata$Age[shotgun_metadata$Species == "SB"], na.rm = TRUE),
            max(shotgun_metadata$Age[shotgun_metadata$Species == "SB"], na.rm = TRUE),
            length.out = 100), Sex = "F",  Location_Type  = "Naïve social group")

pred <- predict(sh1, newdata = newdat3, se.fit = TRUE)
shotgun_metadata$Location_Type
newdat3$fit <- pred$fit
newdat3$upper <- pred$fit + 1.96 * pred$se.fit
newdat3$lower <- pred$fit - 1.96 * pred$se.fit

fig3.b.1 <- ggplot() + geom_point(data = shotgun_metadata[shotgun_metadata$Species == "SB" & shotgun_metadata$Location_Type == "Naïve social group",], 
                                  aes(x = Age, y = shannon, color = boliv_color), alpha = 0.4)  +
  geom_line(data = newdat3, aes(x = Age, y = fit, color = boliv_color), size = 2, col = "black") + ylim(c(4.5,6.5)) +
  geom_ribbon(data = newdat3, aes(x = Age, ymin = lower, ymax = upper, color = boliv_color), alpha = 0.3, color = "black", linewidth = 0.2) +
  theme_pubr(base_size = 24) +
  labs(x = " ", y = "Shannon diversity \n of functional pathways",  title = "Naive social groups") +
  scale_color_manual(values = boliv_color, name = NULL)  + guides(color = "none") 

fig3.b.1 
colnames(shotgun_metadata)
#do the same for number of obs pathways
m_linear <- gam(observed_pathways_new ~ Species + Age + Sex + enclosure + Location_Type,data = shotgun_metadata,method = "REML")
m_smooth <- gam(observed_pathways_new ~ Species + s(Age) + Sex + enclosure + Location_Type, data = shotgun_metadata, method = "REML")

anova(m_linear, m_smooth, test = "F")

AIC(m_linear, m_smooth)#, Species, bs = "fs"

obs_path4 <- gam(observed_pathways_new ~ s(Age)  , data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")
obs_path5 <- gam(observed_pathways_new ~ s(Age)  + Sex  + Location_Type, data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")
obs_path6 <- gam(observed_pathways_new ~ s(Age)  + Sex + enclosure + Location_Type  , data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")
obs_path7 <- gam(observed_pathways_new ~ s(Age)  + Sex + enclosure , data = shotgun_metadata[shotgun_metadata$Species == "SB",], method = "ML")

bbmle::AICctab(obs_path4, obs_path5, obs_path6, obs_path7, weights = T, base = T)

summary(obs_path6)

newdat4 <- expand.grid(Age = seq(min(shotgun_metadata$Age[shotgun_metadata$Species == "SB"], na.rm = TRUE),
            max(shotgun_metadata$Age[shotgun_metadata$Species == "SB"], na.rm = TRUE),
            length.out = 100), Sex = "F", Location_Type = "Naïve social group", enclosure = "Large")

pred <- predict(obs_path6, newdata = newdat4, se.fit = TRUE)

newdat4$fit   <- pred$fit
newdat4$upper <- pred$fit + 1.96 * pred$se.fit
newdat4$lower <- pred$fit - 1.96 * pred$se.fit


fig3.b.2 <- ggplot() + geom_point(data = shotgun_metadata[shotgun_metadata$Species == "SB" & shotgun_metadata$Location_Type == "Naïve social group" & shotgun_metadata$enclosure == "Large",], aes(x = Age, y = observed_pathways_new, color = boliv_color), alpha = 0.4) +
  geom_line(data = newdat4, aes(x = Age, y = fit, color = boliv_color),size = 2, color = "black") +
  geom_ribbon(data = newdat4, aes(x = Age, ymin = lower, ymax = upper, color = boliv_color), alpha = 0.3, color = "black", linewidth = 0.2) +
  theme_pubr(base_size = 24) +  labs(x = "Age (years)", y = "Observed \n functional pathways", title = "Naive social groups") +  scale_color_manual(values = boliv_color, name = NULL) +
  guides(color = "none") + ylim(c(100,500))

fig3.b.2

fig3.b.1 / fig3.b.2

#next set of functional plots

#change reference to treatment group next to see predicted plot
newdat3 <- expand.grid( Age = seq(min(shotgun_metadata$Age[shotgun_metadata$Species == "SB"], na.rm = TRUE),
            max(shotgun_metadata$Age[shotgun_metadata$Species == "SB"], na.rm = TRUE),
            length.out = 100), Sex = "F",  Location_Type  = "Treatment group")

pred <- predict(gam_model1, newdata = newdat3, se.fit = TRUE)

newdat3$fit <- pred$fit
newdat3$upper <- pred$fit + 1.96 * pred$se.fit
newdat3$lower <- pred$fit - 1.96 * pred$se.fit

fig3.b.12 <- ggplot() + geom_point(data = shotgun_metadata[shotgun_metadata$Species == "SB" & shotgun_metadata$Location_Type == "Treatment group",], aes(x = Age, y = shannon, color = boliv_color), alpha = 0.4)  +
  geom_line(data = newdat3, aes(x = Age, y = fit, color = boliv_color), size = 2, col = "black") + ylim(c(4.5,6.5)) +
  geom_ribbon(data = newdat3, aes(x = Age, ymin = lower, ymax = upper, color = boliv_color),
    alpha = 0.3, color = "black", linewidth = 0.2) + theme_pubr(base_size = 24) +
  labs(x = " ", y = "Shannon diversity \n of functional pathways", title = "Treatment group") +
  scale_color_manual(values = boliv_color, name = NULL)  + guides(color = "none") 

fig3.b.12

newdat4 <- expand.grid(Age = seq(min(shotgun_metadata$Age[shotgun_metadata$Species == "SB"], na.rm = TRUE),
            max(shotgun_metadata$Age[shotgun_metadata$Species == "SB"], na.rm = TRUE),
            length.out = 100), Sex = "F", Location_Type = "Treatment group", enclosure = "Large")

pred <- predict(gam_model6, newdata = newdat4, se.fit = TRUE)

newdat4$fit   <- pred$fit
newdat4$upper <- pred$fit + 1.96 * pred$se.fit
newdat4$lower <- pred$fit - 1.96 * pred$se.fit


fig3.b.13 <- ggplot() + geom_point(data = shotgun_metadata[shotgun_metadata$Species == "SB" & shotgun_metadata$Location_Type == "Treatment group",], aes(x = Age, y = observed_pathways_new, color = boliv_color), alpha = 0.4) +
  geom_line(data = newdat4, aes(x = Age, y = fit, color = boliv_color),size = 2, color = "black") +
  geom_ribbon(data = newdat4, aes(x = Age, ymin = lower, ymax = upper, color = boliv_color), alpha = 0.3, color = "black", linewidth = 0.2) +
  theme_pubr(base_size = 24) +  labs(x = "Age (years)", y = "Observed \n functional pathways", title = "Treatment group") +  scale_color_manual(values = boliv_color, name = NULL) +
  guides(color = "none") + ylim(c(100,500))

fig3.b.13


fig3.b.12 / fig3.b.13

#Relative abundances of butyrate, lactate and carbohydrate degrading taxa across age in Sb 
df_genus_summ <- readRDS("sbol_genus_summ.RDS")
head(df_genus_summ)
table(df_genus_summ$Genus, df_genus_summ$sample.id)
str(df_genus_summ)

#make plots of genus categories across age
summary(lm(df_genus_summ$Sum_abund[df_genus_summ$genus_cat == "Butyrate producers"& df_genus_summ$Species == "SB"] ~ df_genus_summ$Age[df_genus_summ$genus_cat == "Butyrate producers"& df_genus_summ$Species == "SB"]))
     
fig3.a <- df_genus_summ %>% filter(genus_cat == "Butyrate producers" & Species == "SB") %>% ggplot(aes(x = Age, y = log1p(Sum_abund + 1e-6), color = Species)) + geom_point(size = 2, alpha = 0.62) + geom_smooth(method = "lm", color = "gray2") + scale_color_manual(values = boliv_color, name = NULL) + theme_pubr(base_size = 24)  + theme(legend.position = "none") + ylab("Relative abundance") +  ggtitle("butyrate producing taxa")

summary(lm(df_genus_summ$Sum_abund[df_genus_summ$genus_cat == "Lactate producers"& df_genus_summ$Species == "SB"] ~ df_genus_summ$Age[df_genus_summ$genus_cat == "Lactate producers"& df_genus_summ$Species == "SB"]))
           
fig3.b <- df_genus_summ %>% filter(genus_cat == "Lactate producers"& Species == "SB") %>% ggplot(aes(x = Age, y = log1p(Sum_abund + 1e-6), color = Species)) + geom_point(size = 2, alpha = 0.62) + geom_smooth(method = "lm", color = "gray2") + scale_color_manual(values = boliv_color, name = NULL) + theme_pubr(base_size = 24)  + theme(legend.position = "none") + ylab(" ") + ggtitle("lactate producing taxa")

confint(lm(df_genus_summ$Sum_abund[df_genus_summ$genus_cat == "Carbohydrate degraders"& df_genus_summ$Species == "SB"] ~ df_genus_summ$Age[df_genus_summ$genus_cat == "Carbohydrate degraders"& df_genus_summ$Species == "SB"]))

fig3.c <- df_genus_summ %>% filter(genus_cat == "Carbohydrate degraders" & Species == "SB") %>% ggplot(aes(x = Age, y = log1p(Sum_abund + 1e-6), color = Species)) + geom_point(size = 2, alpha = 0.62) + geom_smooth(method = "lm", color = "gray2") + scale_color_manual(values = boliv_color, name = NULL) + theme_pubr(base_size = 24)  + theme(legend.position = "none") + ylab("Relative abundance") + ylab("  ") + ggtitle("carbohydrate degrading taxa")

benefic_microbes <- plot_grid(fig3.a, fig3.b, fig3.c, ncol = 3)

ggsave("fig6C_subplot_benefic_microbes.jpeg", benefic_microbes,  height = 8, width= 19, dpi = 450)
```

#7. Differential abundance of pathways in Sb only 
```{r}
boliviensis_pathway_metadata <- readRDS("sbol_pathway_metadata_maaslinSBonly.RDS")
boliviensis_pathway_metadata <- boliviensis_pathway_metadata %>% filter(Location_Type != "Individual") %>% as.data.frame()
table(boliviensis_pathway_metadata$Location_Type) 

#age, sex, loc and enclosure only
fit_funct_boliviensis <- maaslin3(input_data = filtered_pathway,
                                          input_metadata = boliviensis_pathway_metadata,
                                          output = "maaslin3_pathway_SB_bothabunprev_all_Aug11",
                                          formula  = "~ Age + Sex + Location_Type + enclosure", 
                                          normalization = "TSS", 
                                          transform = "LOG",
                                          standardize = FALSE, 
                                          max_significance = 0.1,
                                          max_pngs = 100,
                                          cores = 1, 
                                          save_models = TRUE)


results_path          <- read.delim("maaslin3_pathway_SB_bothabunprev_all_Aug11/all_results.tsv") 
sb_abundance_results_func <- subset(results_path, model == "abundance")
sb_abundance_results_func$log_qval_individual <- -log10(sb_abundance_results_func$qval_individual)
sb_abundance_results_func$color <- ifelse(sb_abundance_results_func$coef < 0 , "lower with age", "higher with age")
#Optional: keep significant hits only
sb_sig_results_funct      <- sb_abundance_results_func %>% filter(qval_joint < 0.05)
head(sb_sig_results_funct)
sort(sb_sig_results_funct$feature)
#rename pathways
sb_sig_results_funct$PathwayID   <- str_remove(sb_sig_results_funct$feature, ":.*")
sb_sig_results_funct$PathwayName <- pathway_name_map[sb_sig_results_funct$PathwayID]

sb_sig_results_funct %>% filter(metadata == "Location_Type" & !is.na(qval_individual) & qval_individual < 0.05 ) %>% nrow() #4
sb_sig_results_funct %>% filter(metadata == "Age" & !is.na(qval_individual) & qval_individual < 0.05  ) %>% nrow() #73
sb_sig_results_funct %>% filter(metadata == "enclosure" & !is.na(qval_individual) & qval_individual < 0.05  ) %>% nrow() #49

sb_sig_results_funct_age         <- sb_sig_results_funct %>% filter(metadata == "Age" & !is.na(qval_individual) & qval_individual < 0.05) %>% as.data.frame()  
sb_sig_results_funct_loc         <- sb_sig_results_funct %>% filter(metadata == "Location_Type" & !is.na(qval_individual) & qval_individual < 0.05) %>% as.data.frame() #3 pathways

#Reorder by coefficient size
sb_sig_results_funct_age$color    <- ifelse(sb_sig_results_funct_age$coef < 0, "lower with age", "higher with age")
sb_sig_results_funct_age          <- sb_sig_results_funct_age %>% arrange(coef) %>% as.data.frame()
sb_sig_results_funct_age$color    <- factor(sb_sig_results_funct_age$color, level = c("lower with age", "higher with age"))


sb_sig_results_funct %>% filter(metadata == "Age" & !is.na(qval_individual) & qval_individual < 0.05  & coef < 0) %>% nrow() #3
sb_sig_results_funct %>% filter(metadata == "Age" & !is.na(qval_individual) & qval_individual < 0.05  & coef > 0) %>% nrow() #70


#Plot only sig # do separate for  "Age"
sb_sig_results_funct_age_fig <- sb_sig_results_funct_age %>% filter(!is.na(qval_individual) & qval_individual < 0.05)  %>%  
  ggplot(aes(x = coef, y = reorder(PathwayName, coef), fill = color)) + 
  geom_col() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("coral","lightblue"),
    name = "direction of effect") + 
  labs(title = "Differentially abundant functional pathways with age in SB",
    x = "Model coefficient (beta)",
    y = "Pathway") +  theme_pubr(base_size = 22)  + theme(panel.grid.major.y = element_blank(), legend.position = "right")

ggsave( "BMC_BioFig7.jpeg", sb_sig_results_funct_age_fig, width = 18, height = 24, units = "in", dpi = 440) 

sb_sig_age_table <- sb_sig_results_funct_age %>% as.data.frame()
write.csv(sb_sig_age_table, "maaslin3_sig_SB_age_table.csv", row.names = F)

sb_abundance_results_func %>% filter(metadata == "Age" & !is.na(coef)) %>% ggplot(aes(x = coef, y = log_qval_individual, col = color)) +
geom_jitter(size = 4, alpha = 0.75) + geom_hline(yintercept = 1.3, linetype = "dotted", color = "maroon") +
  geom_vline(xintercept = 0,
             linetype = "dashed") +  scale_fill_manual(values = c( "coral","lightblue"),  name = "qval_joint < 0.05") + 
  theme_pubr(base_size = 24) +
  labs( x = "Beta coefficient",
    y = "-log10(q-value)",
    title = "functional pathway abundance associations \n in SB with age") + theme(plot.title = element_text(hjust = 0.5))
#how many higher or lower?

abundance_results_func %>% filter(metadata == "Age" & !is.na(qval_individual) & qval_individual < 0.05 & coef > 0) %>% nrow() #49 
abundance_results_func %>% filter(metadata == "Age" & !is.na(qval_individual) & qval_individual < 0.05 & coef < 0) %>% nrow() #2

#plot the max ones? NO
abundance_results_func %>% filter(metadata == "Age" & !is.na(qval_individual) & qval_individual < 0.05 ) %>% arrange(coef) #2


##genus and SCFA plot
genus_wide <- df_genus_summ %>%
  dplyr::select(sample.id, Genus, Sum_abund) %>%
  distinct() %>%
  pivot_wider(names_from = Genus, values_from = Sum_abund, values_fill = 0) %>% as.data.frame()

head(genus_wide)
scfa_data <- df_genus_summ %>%
  dplyr::select(sample.id, acetate, propionate, butyrate, isovalerate) %>%
  distinct() %>% as.data.frame()

genus_wide <- genus_wide %>% arrange(sample.id)
scfa_data  <- scfa_data  %>% arrange(sample.id)
all(genus_wide$sample.id == scfa_data$sample.id)

#prepare matrices
genus_mat <- genus_wide %>% dplyr::select(-sample.id)
scfa_mat  <- scfa_data %>% dplyr::select(-sample.id)
genus_mat <- log1p(genus_mat)

cor_mat    <- cor(genus_mat, scfa_mat, method = "spearman", use = "pairwise.complete.obs")
scfa_cols  <- c("acetate", "propionate", "butyrate", "isovalerate")
genus_cols <- colnames(genus_mat)

library(ComplexHeatmap)
library(circlize)
library(Hmisc)
library(pheatmap)
res <- rcorr(as.matrix(genus_mat), as.matrix(scfa_mat), type = "spearman")

p_mat <- res$P[genus_cols, scfa_cols]
p_adj <- matrix(p.adjust(p_mat, method = "fdr"), nrow = nrow(p_mat),dimnames = dimnames(p_mat))
stars <- ifelse(p_adj < 0.001, "***",
         ifelse(p_adj < 0.01,  "**",
         ifelse(p_adj < 0.05,  "*", "")))

pheatmap(cor_mat,  col = colorRampPalette(c("coral3", "white", "cyan3"))(100), cluster_rows = TRUE, cluster_cols  = FALSE, 
        display_numbers = stars, number_color = "black", 
         fontsize = 15 , fontsize_col = 18, fontsize_number = 8) #change this as per genus type


row_ha <- rowAnnotation(
  Species = df$Species,
  col = list(
    Species = c(
      "Species1" = "#1b9e77",
      "Species2" = "#d95f02",
      "Species3" = "#7570b3"
    )
  )
)

#add info on genus here 
SCFA_group        <- c("Faecalibacterium", "Anaerostipes", "Agathobaculum", "Ruminococcus","Gemmiger", "Fusicatenibacter")
Facultative_group <- c("Streptococcus","Lactobacillus","Lactococcus","Staphylococcus","Moraxella")

#Microbiome Quality Score = (Sum of SCFA producers) / (Sum of facultative/aerotolerant taxa)

df_genus_SCFA           <- df_genus_summ %>% group_by(sample.id) %>% 
                           filter(Genus %in% SCFA_group) %>% mutate(sum_of_SCFA = sum(Sum_abund)) %>% as.data.frame()

df_genus_facultative    <- df_genus_summ %>% group_by(sample.id) %>% 
                           filter(Genus %in% Facultative_group) %>% mutate(sum_of_facultative = sum(Sum_abund)) %>% as.data.frame() 

#calculate the ratios in each sample 
#proportion of SCFA producing genus 
df_genus$sum_of_SCFA        <- ifelse(df_genus$Genus %in% SCFA_group, df_genus_SCFA$sum_of_SCFA[match(df_genus$sample.id,df_genus_SCFA$sample.id)], NA )
df_genus$sum_of_facultative <- ifelse(df_genus$Genus %in% Facultative_group, df_genus_facultative$sum_of_facultative[match(df_genus$sample.id,df_genus_facultative$sample.id)], NA )

df_genus$prop_SCFA         <- df_genus$Abundance/df_genus$sum_of_SCFA
df_genus$prop_facultative  <- df_genus$Abundance/df_genus$sum_of_facultative
df_genus$age_group         <- meta$group[match(df_genus$sample.id, meta$sample.id)]
library(ggplot2)

df_genus %>% filter(!is.na(prop_SCFA) & Abundance != 0) %>% ggplot(aes(x = age_group, y = prop_SCFA, col = age_group)) +
  geom_violin() +
  labs(title = "Beneficial Microbe Ratio Across Age Groups",
       x = "Age Group",
       y = "Ratio") + facet_grid(.~Species) +
  ggpubr::theme_pubr(base_size = 18, base_family = "Avenir") 

df_genus %>% filter(!is.na(prop_facultative) & Abundance != 0) %>% ggplot(aes(x = age_group, y = prop_facultative, col = age_group)) +
  geom_boxplot() +
  labs(title = "Facultative Microbe Ratio Across Age Groups",
       x = "Age Group",
       y = "Ratio") + facet_grid(.~Species) +
  ggpubr::theme_pubr(base_size = 18, base_family = "Avenir") 

anova_res <- aov(prop_facultative ~ age_group, data = df_genus)
summary(anova_res)
kruskal.test(ratio ~ age_group, data = df) 

#pathogenic taxa - check 

```

#8.GBMs in Sb only across age
```{r}
#read in GBM map data
func_mat_gbm           <- readRDS("sbol_gbm_data_for_maaslin3_Aug10.rds")
maaslin_metadata       <- readRDS( "sbol_GBM_metadata_maaslin_Aug112926.RDS")
head(maaslin_metadata)
str(maaslin_metadata)
table(maaslin_metadata$Sample)

#differntial abundance with everything
maaslin_output   <- maaslin3(input_data = func_mat_gbm,
                           input_metadata = boliviensis_pathway_metadata,
                           output = 'maaslin_GBM_Aug10_all_covariates_prevabund_SBonly', 
                           formula = '~ Sex + Age + enclosure + Location_Type + log_reads', 
                           normalization = 'TSS',
                           transform = 'LOG',
                           max_significance = 0.1, 
                           max_pngs = 100,
                           cores = 1,
                           save_models = TRUE) 

sig_gbms <- fread("maaslin_GBM_Aug10_all_covariates_prevabund_SBonly/significant_results.tsv")
"Glutamate synthesis I"
"Tryptophan synthesis" 
"Acetate synthesis I/II/III"
"Menaquinone synthesis (vitamin K2) II"

#age specific features
age_feat <- sig_gbms$feature[sig_gbms$metadata == "Age"]
age_feat
length(age_feat)
age_gbm <- sig_gbms %>% filter(metadata == "Age" & !is.na(qval_individual) & qval_individual < 0.05 ) %>% as.data.frame() 

sig_gbms %>% filter(metadata == "Age" & !is.na(qval_individual) & qval_individual < 0.05 & coef < 0 ) %>% as.data.frame() %>% nrow() #1 decrese with age
sig_gbms %>% filter(metadata == "Age" & !is.na(qval_individual) & qval_individual < 0.05 & coef > 0 ) %>% as.data.frame() %>% nrow() #7 increase with age

#pathway of interest
pos                <- which(colnames(func_mat) == age_feat[1])
pathway_tss <- func_tss[, pos]

pseudo   <- min(pathway_tss[pathway_tss > 0], na.rm = TRUE) / 2

pathway_log2       <- log2(pathway_tss + pseudo)
values             <- pathway_log2

#create a plot for each of these features now
sampleid           <- rownames(func_mat)
temp_data          <- as.data.frame(cbind(sampleid, values))
temp_data$species  <- maaslin_metadata$Species[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$values   <- as.numeric(temp_data$values)
temp_data$age      <- maaslin_metadata$Age[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$sex      <- maaslin_metadata$Sex[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group<- maaslin_metadata$Age_group1[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group2 <- maaslin_metadata$Age_Group[match(temp_data$sampleid, maaslin_metadata$Sample)]

gbm      <- lm(values ~ species + age, data = temp_data)
summary(gbm)

sam <- temp_data %>% filter(species == "SB") %>%  ggplot(aes(x = age, y = values, color = species)) + geom_point(size = 4, alpha = 0.6) + geom_smooth(method = "lm") + theme_pubr(base_size = 24) + theme(legend.position = "right") + scale_color_manual(values = boliv_color, name = NULL) + ylab("\n pathway abundance \n  (TSS normalized:log2 transformed)") + xlab("")  +  ggtitle(age_feat[1])  + xlab("Age in years") + theme(legend.position = "none")

#pathway of interest
pos                <- which(colnames(func_mat) == age_feat[2])
pathway_tss <- func_tss[, pos]

pseudo   <- min(pathway_tss[pathway_tss > 0], na.rm = TRUE) / 2

pathway_log2       <- log2(pathway_tss + pseudo)
values             <- pathway_log2

#create a plot for each of these features now
sampleid           <- rownames(func_mat)
temp_data          <- as.data.frame(cbind(sampleid, values))
temp_data$species  <- maaslin_metadata$Species[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$values   <- as.numeric(temp_data$values)
temp_data$age      <- maaslin_metadata$Age[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$sex      <- maaslin_metadata$Sex[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group<- maaslin_metadata$Age_group1[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group2 <- maaslin_metadata$Age_Group[match(temp_data$sampleid, maaslin_metadata$Sample)]

gbm      <- lm(values ~ species + age, data = temp_data)
summary(gbm)

tryp <- temp_data %>% filter(species == "SB") %>%  ggplot(aes(x = age, y = values, color = species)) + geom_point(size = 4, alpha = 0.6) + geom_smooth(method = "lm") + theme_pubr(base_size = 24) + theme(legend.position = "right") + scale_color_manual(values = boliv_color, name = NULL) + ylab("\n pathway abundance \n  (TSS normalized:log2 transformed)") + xlab("")  +  ggtitle(age_feat[2])  + xlab("Age in years") + theme(legend.position = "none")

#pathway of interest
pos                <- which(colnames(func_mat) == age_feat[3])
pathway_tss <- func_tss[, pos]

pseudo   <- min(pathway_tss[pathway_tss > 0], na.rm = TRUE) / 2

pathway_log2       <- log2(pathway_tss + pseudo)
values             <- pathway_log2

#create a plot for each of these features now
sampleid           <- rownames(func_mat)
temp_data          <- as.data.frame(cbind(sampleid, values))
temp_data$species  <- maaslin_metadata$Species[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$values   <- as.numeric(temp_data$values)
temp_data$age      <- maaslin_metadata$Age[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$sex      <- maaslin_metadata$Sex[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group<- maaslin_metadata$Age_group1[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group2 <- maaslin_metadata$Age_Group[match(temp_data$sampleid, maaslin_metadata$Sample)]

gbm      <- lm(values ~ species + age, data = temp_data)
summary(gbm)

acet <- temp_data %>% filter(species == "SB") %>%  ggplot(aes(x = age, y = values, color = species)) + geom_point(size = 4, alpha = 0.6) + geom_smooth(method = "lm") + theme_pubr(base_size = 24) + theme(legend.position = "right") + scale_color_manual(values = boliv_color, name = NULL) + ylab("\n pathway abundance \n  (TSS normalized:log2 transformed)") + xlab("")  +  ggtitle(age_feat[3])  + xlab("Age in years") + theme(legend.position = "none")


#pathway of interest
pos                <- which(colnames(func_mat) == age_feat[4])
pathway_tss <- func_tss[, pos]

pseudo   <- min(pathway_tss[pathway_tss > 0], na.rm = TRUE) / 2

pathway_log2       <- log2(pathway_tss + pseudo)
values             <- pathway_log2

#create a plot for each of these features now
sampleid           <- rownames(func_mat)
temp_data          <- as.data.frame(cbind(sampleid, values))
temp_data$species  <- maaslin_metadata$Species[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$values   <- as.numeric(temp_data$values)
temp_data$age      <- maaslin_metadata$Age[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$sex      <- maaslin_metadata$Sex[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group<- maaslin_metadata$Age_group1[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group2 <- maaslin_metadata$Age_Group[match(temp_data$sampleid, maaslin_metadata$Sample)]

gbm      <- lm(values ~ species + age, data = temp_data)
summary(gbm)

acet2 <- temp_data %>% filter(species == "SB") %>%  ggplot(aes(x = age, y = values, color = species)) + geom_point(size = 4, alpha = 0.6) + geom_smooth(method = "lm") + theme_pubr(base_size = 24) + theme(legend.position = "right") + scale_color_manual(values = boliv_color, name = NULL) + ylab("\n pathway abundance \n  (TSS normalized:log2 transformed)") + xlab("")  +  ggtitle(age_feat[4])  + xlab("Age in years") + theme(legend.position = "none")

#pathway of interest
pos                <- which(colnames(func_mat) == age_feat[5])
pathway_tss <- func_tss[, pos]

pseudo   <- min(pathway_tss[pathway_tss > 0], na.rm = TRUE) / 2

pathway_log2       <- log2(pathway_tss + pseudo)
values             <- pathway_log2

#create a plot for each of these features now
sampleid           <- rownames(func_mat)
temp_data          <- as.data.frame(cbind(sampleid, values))
temp_data$species  <- maaslin_metadata$Species[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$values   <- as.numeric(temp_data$values)
temp_data$age      <- maaslin_metadata$Age[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$sex      <- maaslin_metadata$Sex[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group<- maaslin_metadata$Age_group1[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group2 <- maaslin_metadata$Age_Group[match(temp_data$sampleid, maaslin_metadata$Sample)]

gbm      <- lm(values ~ species + age, data = temp_data)
summary(gbm)

glut <- temp_data %>% filter(species == "SB") %>%  ggplot(aes(x = age, y = values, color = species)) + geom_point(size = 4, alpha = 0.6) + geom_smooth(method = "lm") + theme_pubr(base_size = 24) + theme(legend.position = "right") + scale_color_manual(values = boliv_color, name = NULL) + ylab("\n pathway abundance \n  (TSS normalized:log2 transformed)") + xlab("")  +  ggtitle(age_feat[5])  + xlab("Age in years") + theme(legend.position = "none")

#pathway of interest
pos                <- which(colnames(func_mat) == age_feat[6])
pathway_tss <- func_tss[, pos]

pseudo   <- min(pathway_tss[pathway_tss > 0], na.rm = TRUE) / 2

pathway_log2       <- log2(pathway_tss + pseudo)
values             <- pathway_log2

#create a plot for each of these features now
sampleid           <- rownames(func_mat)
temp_data          <- as.data.frame(cbind(sampleid, values))
temp_data$species  <- maaslin_metadata$Species[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$values   <- as.numeric(temp_data$values)
temp_data$age      <- maaslin_metadata$Age[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$sex      <- maaslin_metadata$Sex[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group<- maaslin_metadata$Age_group1[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group2 <- maaslin_metadata$Age_Group[match(temp_data$sampleid, maaslin_metadata$Sample)]

gbm      <- lm(values ~ species + age, data = temp_data)
summary(gbm)

ino <- temp_data %>% filter(species == "SB") %>%  ggplot(aes(x = age, y = values, color = species)) + geom_point(size = 4, alpha = 0.6) + geom_smooth(method = "lm") + theme_pubr(base_size = 24) + theme(legend.position = "right") + scale_color_manual(values = boliv_color, name = NULL) + ylab("\n pathway abundance \n  (TSS normalized:log2 transformed)") + xlab("")  +  ggtitle(age_feat[6])  + xlab("Age in years") + theme(legend.position = "none")

#pathway of interest
pos                <- which(colnames(func_mat) == age_feat[7])
pathway_tss <- func_tss[, pos]

pseudo   <- min(pathway_tss[pathway_tss > 0], na.rm = TRUE) / 2

pathway_log2       <- log2(pathway_tss + pseudo)
values             <- pathway_log2

#create a plot for each of these features now
sampleid           <- rownames(func_mat)
temp_data          <- as.data.frame(cbind(sampleid, values))
temp_data$species  <- maaslin_metadata$Species[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$values   <- as.numeric(temp_data$values)
temp_data$age      <- maaslin_metadata$Age[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$sex      <- maaslin_metadata$Sex[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group<- maaslin_metadata$Age_group1[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group2 <- maaslin_metadata$Age_Group[match(temp_data$sampleid, maaslin_metadata$Sample)]

gbm      <- lm(values ~ species + age, data = temp_data)
summary(gbm)

vitk <- temp_data %>% filter(species == "SB") %>%  ggplot(aes(x = age, y = values, color = species)) + geom_point(size = 4, alpha = 0.6) + geom_smooth(method = "lm") + theme_pubr(base_size = 24) + theme(legend.position = "right") + scale_color_manual(values = boliv_color, name = NULL) + ylab("\n pathway abundance \n  (TSS normalized:log2 transformed)") + xlab("")  +  ggtitle(age_feat[7])  + xlab("Age in years") + theme(legend.position = "none")


#pathway of interest
pos                <- which(colnames(func_mat) == age_feat[8])
pathway_tss <- func_tss[, pos]

pseudo   <- min(pathway_tss[pathway_tss > 0], na.rm = TRUE) / 2

pathway_log2       <- log2(pathway_tss + pseudo)
values             <- pathway_log2

#create a plot for each of these features now
sampleid           <- rownames(func_mat)
temp_data          <- as.data.frame(cbind(sampleid, values))
temp_data$species  <- maaslin_metadata$Species[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$values   <- as.numeric(temp_data$values)
temp_data$age      <- maaslin_metadata$Age[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$sex      <- maaslin_metadata$Sex[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group<- maaslin_metadata$Age_group1[match(temp_data$sampleid, maaslin_metadata$Sample)]
temp_data$age_group2 <- maaslin_metadata$Age_Group[match(temp_data$sampleid, maaslin_metadata$Sample)]

gbm      <- lm(values ~ species + age, data = temp_data)
summary(gbm)

acet3 <- temp_data %>% filter(species == "SB") %>%  ggplot(aes(x = age, y = values, color = species)) + geom_point(size = 4, alpha = 0.6) + geom_smooth(method = "lm") + theme_pubr(base_size = 24) + theme(legend.position = "right") + scale_color_manual(values = boliv_color, name = NULL) + ylab("\n pathway abundance \n  (TSS normalized:log2 transformed)") + xlab("")  +  ggtitle(age_feat[8])  + xlab("Age in years") + theme(legend.position = "none")

age_gbm_fig <- plot_grid(acet, tryp, glut, sam, ino, vitk, ncol = 3, nrow = 2)

jpeg("BMC_BioFig8.jpeg", width = 70, height = 34, res = 450, units = "cm")
age_gbm_fig
dev.off()
#which are the ones differentially abundant 
sig_GBMs <- read.csv("/Users/sharmisen/Desktop/POSTDOC PROJECTS/NORTHWESTERN/Squirrel Monkey Project/GBM_pathways_associations.csv")
head(sig_GBMs)
#subset to those only
func_mat <- func_mat[func_mat$ModuleName %in% sig_GBMs$feature,]

table(colnames(func_mat))
which(func_mat_t$sample.id == 6524) #6524 is repeated twice - remove 
func_mat_t               <- func_mat_t[-c(69,99),] 
rownames(func_mat_t)     <- func_mat_t$sample.id
func_mat_t               <- func_mat_t %>% dplyr::select(-sample.id)
func_mat_t_num           <- apply(func_mat_t, 2, as.numeric)
rownames(func_mat_t_num) <- rownames(func_mat_t)
func_clr <- clr(func_mat_t_num + 1e-6)   # transform abundances
which(rownames(func_clr) %!in% meta_func$Animal)
meta_func_ordered <- meta_func[match(as.numeric(rownames(func_clr)),meta_func$Animal),]

surv_obj <- Surv(time = meta_func_ordered$time, event = meta_func_ordered$status)

#univariate cox model for each functional pathway
results <- data.frame()

for (feature in colnames(func_clr)) {
  values <- as.numeric(func_mat_t[, feature] )
  model <- coxph(surv_obj ~ values + Age_group1 + Sex + Species ,
                 data = meta_func_ordered)
  s <- summary(model)
  results <- rbind(results, data.frame(
    feature = feature,
    HR = s$coef[1, "exp(coef)"],
    pval = s$coef[1, "Pr(>|z|)"]
  ))
}

results$padj <- p.adjust(results$pval, method = "fdr")

 
pca <- prcomp(func_clr, scale. = TRUE)

meta_func_ordered$PC1 <- pca$x[,1]
meta_func_ordered$PC2 <- pca$x[,2]
surv_obj <- Surv(time = meta_func_ordered$time, event = meta_func_ordered$status)

cox_model <- coxph(surv_obj ~ PC1 + PC2 + Age_group1 + Sex * Species ,
                   data = meta_func_ordered)
summary(cox_model)

library(survminer)

#example: split PC1 at median
meta_func_ordered$PC1_group <- ifelse(meta_func_ordered$PC1 > median(meta_func_ordered$PC1), "High", "Low")
meta_func_ordered$PC2_group <- ifelse(meta_func_ordered$PC2 > median(meta_func_ordered$PC1), "High", "Low")

fit <- survfit(surv_obj ~ PC2_group, data = meta_func_ordered)

ggsurvplot(fit, data = meta_func_ordered, pval = TRUE)
meta_func_ordered_adults <- meta_func_ordered %>% filter(Age_group1 %in% c("Adult", "Geriatric", "Geriatric_old")) %>% as.data.frame()

surv_obj <- Surv(time = meta_func_ordered_adults$time, event = meta_func_ordered_adults$status)

fit      <- survfit(surv_obj ~ Age_group1, data = meta_func_ordered_adults)
summary(fit)
p <- ggsurvplot(fit, data = meta_func_ordered_adults, pval = TRUE,
                palette = c("#CC79A7","purple3" ,"gray"), font.legend = 14, 
                legend.position = "none",
                size = 1.5)
p 

fit      <- survfit(surv_obj ~ Species, data = meta_func_ordered_adults)

p1 <- ggsurvplot(fit, data = meta_func_ordered_adults, pval = TRUE, 
                 palette = c("yellow3","purple3"), font.legend = 14, legend.position = "none",
                size = 1.5) 

p1 

```
#9. SCFA models
```{r}
targeted_data <- shotgun_metadata
#modeling the SCFA data
head(targeted_data)
#remove hospital and recovery from the targeted data 
full_targeted_data <- targeted_data %>% filter(Location_Type != "Individual") %>% 
                                        dplyr::select(Age , Sex, Species, Location_Type, enclosure, propionate, acetate, butyrate, valerate, total_SCFA) %>% 
                                        mutate(prop_PA = propionate/total_SCFA, prop_BA = butyrate/total_SCFA, 
                                               prop_AA = acetate/ total_SCFA, prop_VA = valerate/total_SCFA) %>% as.data.frame()

#scale and standardize age_cont and weight variables 
full_targeted_data$age_scaled  <- scale(full_targeted_data$Age)
md.pattern(full_targeted_data)
complete_targeted_data <- cc(full_targeted_data)
complete_targeted_data <- complete_targeted_data %>% filter(Species == "SB") %>% as.data.frame()
#modelling each SCFA metabolite separately without any other metabolite present

#PA - propionate - setup a subset of models 
pa.models0 <- glm(prop_PA ~ 1 , data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
pa.models1 <- glm(prop_PA ~ enclosure, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
pa.models2 <- glm(prop_PA ~ age_scaled, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
pa.models3 <- glm(prop_PA ~ age_scaled + Location_Type, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
pa.models4 <- glm(prop_PA ~ age_scaled + Sex, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
pa.models5 <- glm(prop_PA ~ age_scaled + Sex  + Location_Type , data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
pa.models6 <- glm(prop_PA ~ age_scaled + Sex  + Location_Type + enclosure , data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
pa.models7 <- glm(prop_PA ~ age_scaled * Sex  + Location_Type , data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail") 
pa.models8 <- glm(prop_PA ~ age_scaled * Sex  + Location_Type + enclosure, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")

#Put all models in a list
pa_models <- list(pa.models0, pa.models1, pa.models2, pa.models3, pa.models4,pa.models5, pa.models6, pa.models7, pa.models8)
#adding model names for clarity in the output
model_names <- c("pa.models0","pa.models1", "pa.models2", "pa.models3", "pa.models4","pa.models5", "pa.models6", "pa.models7", "pa.models8")

#Run aictab with weights
AICtab(pa.models0, pa.models1, pa.models2, pa.models3, pa.models4,pa.models5, pa.models6, pa.models7, pa.models8, weights = TRUE, base = T)

top_pa_models      <- list(pa.models2, pa.models4)
top_pa_model_names <- c("pa.models2", "pa.models4")

#Model averaging
pa_ma <- model.avg(object =  top_pa_models, model_names = top_pa_model_names, second.ord = TRUE) # uses AICc weights

#View averaged coefficients
summary(pa_ma)

#Full = TRUE includes predictors that aren't in all models
coef_df <- as.data.frame(coef(pa_ma, full = FALSE))
conf_df <- as.data.frame(confint(pa_ma, full = FALSE))

#Combine into a single dataframe
plot_df <- data.frame(Predictor = rownames(coef_df),
  Estimate = coef_df$`coef(pa_ma, full = FALSE)`,
  Lower = conf_df$`2.5 %`,
  Upper = conf_df$`97.5 %`)

# Optional: remove intercept if you want only predictors
plot_df   <- plot_df[plot_df$Predictor != "(Intercept)", ]
plot_df$Predictor <- c("Age", "Male")
prop_plot <- ggplot(plot_df, aes(x = Predictor, y = Estimate)) +
  geom_point(size = 6, color = "black") +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.05, color = "black") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray", linewidth = 2) + 
  coord_flip() +  # flip for horizontal labels 
  labs( y = " ", x = "", title = "Propionate") + theme_pubr(base_size = 24)

prop_plot

#butyrate - setup a subset of models 
ba.models0         <-  glm(prop_BA ~ 1 , data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
ba.models1         <-  glm(prop_BA ~ enclosure, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
ba.models2         <-  glm(prop_BA ~ age_scaled, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
ba.models3         <-  glm(prop_BA ~ age_scaled + Location_Type, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
ba.models4         <-  glm(prop_BA ~ age_scaled + Sex, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
ba.models5         <-  glm(prop_BA ~ age_scaled + Sex  + Location_Type + enclosure, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
ba.models6         <-  glm(prop_BA ~ age_scaled + Sex  + Location_Type , data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
ba.models7         <-  glm(prop_BA ~ age_scaled * Sex  + Location_Type , data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail") 
ba.models8         <-  glm(prop_BA ~ age_scaled * Sex  + Location_Type + enclosure, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")

#Put all models in a list
ba_models          <- list(ba.models0, ba.models1, ba.models2, ba.models3, ba.models4,ba.models5, ba.models6, ba.models7, ba.models8)
#adding model names for clarity in the output
model_names        <- c("ba.models0","ba.models1", "ba.models2", "ba.models3", "ba.models4","ba.models5", "ba.models6", "ba.models7", "ba.models8")

#Run aictab with weights
AICtab(ba.models0, ba.models1, ba.models2, ba.models3, ba.models4,ba.models5, ba.models6, ba.models7, ba.models8,  weights = TRUE, base = T)

top_ba_models       <- list(ba.models4, ba.models2)
top_ba_model_names  <- c("ba.models4","ba.models2")

#Model averaging
ba_ma               <- model.avg(object =  top_ba_models, model_names = top_ba_model_names, second.ord = TRUE) # uses AICc weights

#View averaged coefficients
summary(ba_ma)

#Full = TRUE includes predictors that aren't in all models
coef_df             <- as.data.frame(coef(ba_ma, full = FALSE))
conf_df             <- as.data.frame(confint(ba_ma, full = FALSE))

#Combine into a single dataframe
plot_df             <- data.frame(Predictor = rownames(coef_df), Estimate = coef_df$`coef(ba_ma, full = FALSE)`, Lower = conf_df$`2.5 %`, Upper = conf_df$`97.5 %`)

#Optional: remove intercept if you want only predictors
plot_df           <- plot_df[plot_df$Predictor != "(Intercept)", ]
plot_df$Predictor <- c("Age", "Male")
but_plot          <- ggplot(plot_df, aes(x = Predictor, y = Estimate)) + geom_point(size = 6, color = "black") +
                                     geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.05, color = "black") +
                                     geom_hline(yintercept = 0, linetype = "dashed", color = "gray", linewidth = 2) + 
                                     coord_flip() +  # flip for horizontal labels 
                                     labs( y = " ", x = "", title = "Butyrate") + theme_pubr(base_size = 24)
but_plot

#acetate - setup a subset of models
aa.models0 <- glm(prop_AA ~ 1 , data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
aa.models1 <- glm(prop_AA ~ enclosure, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
aa.models2 <- glm(prop_AA ~ age_scaled, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
aa.models3 <- glm(prop_AA ~ age_scaled + Location_Type, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
aa.models4 <- glm(prop_AA ~ age_scaled +  Sex, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
aa.models5 <- glm(prop_AA ~ age_scaled + Sex  + Location_Type + enclosure, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
aa.models6 <- glm(prop_AA ~ age_scaled + Sex  + Location_Type , data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
aa.models7 <- glm(prop_AA ~ age_scaled * Sex  + enclosure + Location_Type , data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail") 
aa.models8 <- glm(prop_AA ~ age_scaled * Sex  + Location_Type , data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")

#Put all models in a list
aa_models <- list(aa.models0, aa.models1, aa.models2, aa.models3, aa.models4,aa.models5, aa.models6, aa.models7, aa.models8)
#adding model names for clarity in the output
model_names <- c("aa.models0","aa.models1", "aa.models2", "aa.models3", "aa.models4","aa.models5", "aa.models6", "aa.models7", "aa.models8")

#Run aictab with weights
AICtab(aa.models0, aa.models1, aa.models2, aa.models3, aa.models4,aa.models5, aa.models6, aa.models7, aa.models8, weights = TRUE, base = TRUE)

top_aa_models      <- list(aa.models4, aa.models2)
top_aa_model_names <- c("aa.models4","aa.models2")

#Model averaging
aa_ma <- model.avg(object =  top_aa_models, model_names = top_aa_model_names, second.ord = TRUE) # uses AICc weights

#View averaged coefficients
summary(aa_ma)

#Full = TRUE includes predictors that aren't in all models
coef_df <- as.data.frame(coef(aa_ma, full = FALSE))
conf_df <- as.data.frame(confint(aa_ma, full = FALSE))

#Combine into a single dataframe
plot_df <- data.frame(Predictor = rownames(coef_df),
  Estimate = coef_df$`coef(aa_ma, full = FALSE)`,
  Lower = conf_df$`2.5 %`,
  Upper = conf_df$`97.5 %`)

# Optional: remove intercept if you want only predictors
plot_df <- plot_df[plot_df$Predictor != "(Intercept)", ]
plot_df$Predictor <- c("Age", "Male")
acet_plot <- ggplot(plot_df, aes(x = Predictor, y = Estimate)) +
  geom_point(size = 6, color = "black") +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.05, color = "black")  +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray", linewidth = 2) + 
  coord_flip() +  # flip for horizontal labels 
  labs( y = "Model-averaged coefficient", x = "", title = "Acetate") + theme_pubr(base_size = 24)
acet_plot

#isovalerate
#isovalerate - setup a subset of models 
iVa.models0 <- glm(prop_VA ~ 1 , data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
iVa.models1 <- glm(prop_VA ~ enclosure, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
iVa.models2 <- glm(prop_VA ~ age_scaled, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
iVa.models3 <- glm(prop_VA ~ age_scaled + enclosure, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
iVa.models4 <- glm(prop_VA ~ age_scaled + Sex, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
iVa.models5 <- glm(prop_VA ~ age_scaled + Sex + Location_Type , data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
iVa.models6 <- glm(prop_VA ~ age_scaled + Location_Type + enclosure, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
iVa.models7 <- glm(prop_VA ~ age_scaled + Sex + Location_Type + enclosure, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail") 
iVa.models8 <- glm(prop_VA ~ age_scaled * Sex + Location_Type, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
iVa.models9 <- glm(prop_VA ~ age_scaled * Sex + enclosure, data = complete_targeted_data, family = Gamma(link = "log"), na.action = "na.fail")
#Put all models in a list
iVa_models <- list(iVa.models0, iVa.models1, iVa.models2, iVa.models3, iVa.models4,iVa.models5, iVa.models6, iVa.models7, iVa.models8, iVa.models9)
#adding model names for clarity in the output
model_names <- c("iVa.models0","iVa.models1", "iVa.models2", "iVa.models3", "iVa.models4","iVa.models5", "iVa.models6", "iVa.models7", "iVa.models8", "iVa.models9")

#Run aictab with weights
AICctab(iVa.models0, iVa.models1, iVa.models2, iVa.models3, iVa.models4,iVa.models5, iVa.models6, iVa.models7, iVa.models8, iVa.models9, weights = TRUE, base = T)

top_va_models      <- list(iVa.models0, iVa.models4, iVa.models1)
top_va_model_names <- c("iVa.models0","iVa.models4", "iVa.models1")

#Model averaging
iVa_ma <- model.avg(object =  top_va_models, model_names = top_va_model_names, second.ord = TRUE) # uses AICc weights

#View averaged coefficients
summary(iVa_ma)

#Full = TRUE includes predictors that aren't in all models
coef_df <- as.data.frame(coef(iVa_ma, full = FALSE))
conf_df <- as.data.frame(confint(iVa_ma, full = FALSE))

#Combine into a single dataframe
plot_df <- data.frame(Predictor = rownames(coef_df),
  Estimate = coef_df$`coef(iVa_ma, full = FALSE)`,
  Lower = conf_df$`2.5 %`,
  Upper = conf_df$`97.5 %`)

# Optional: remove intercept if you want only predictors
plot_df <- plot_df[plot_df$Predictor != "(Intercept)", ]
plot_df$Predictor <- c("Age", "Male", "Small", "Medium", "Large")
#plot_df <- plot_df[plot_df$Predictor != "Small enclosure",]

iva_plot <- ggplot(plot_df, aes(x = Predictor, y = Estimate)) +
  geom_point(size = 6, color = "black") +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.05, color = "black")  +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray", linewidth = 2) + 
  coord_flip() +  # flip for horizontal labels 
  labs( y = "Model-averaged coefficient", x = "", title = "isoValerate") + theme_pubr(base_size = 24)
iva_plot 

scfa_model_plot <- cowplot::plot_grid(prop_plot, but_plot, iva_plot , acet_plot, ncol = 2) #Fig 9
#ggsave( "SCFA_models.png",scfa_model_plot, width = 18, height = 11, units = "in", dpi = 340)
```

#Differential Metabolites with age in Sb only
```{r}
#do it only for SB
head(metadata)
str(metadata)
metadata$Age <- as.numeric(metadata$Age)
metadata_SB_limma <- metadata %>% filter(Species == "SB") %>% as.data.frame()
#filter by only the SB 
metabolite_matrix_check_SB <- metabolite_matrix_check[,colnames( metabolite_matrix_check) %in% metadata_SB_limma$normalized_name]

design <- model.matrix(~ Age + Sex + Location_Type + enclosure,
  data = metadata_SB_limma)

colnames(metabolite_matrix_check) 

fit2 <- lmFit(metabolite_matrix_check_SB, design)
fit2 <- eBayes(fit2)
dim(fit2)
summary(fit2)
sb_age_res <- topTable(fit2, coef = "Age", number = Inf, adjust.method = "BH")

sb_age_res$significant <- ifelse(sb_age_res$adj.P.Val < 0.05, "significant", "not significant")
volcano_metab <- ggplot(sb_age_res,aes(x = logFC,   y = -log10(adj.P.Val), color = significant)) +
 geom_point(alpha = 0.65, size = 3) + 
  theme_pubr(base_size = 20) + geom_vline(xintercept = 0, linetype = 2) + theme(legend.position = "right") +
  labs(x = "Beta coefficient", y = "-log10(FDR)", title = "untargeted metabolite abundance \n associations (with age)") + theme(plot.title = element_text(hjust = 0.5))

ggsave("volcano_metab.png", volcano_metab, width = 8, height = 6, units = "in", dpi = 320)

report_file_metab_sb <- sb_age_res %>% filter(significant ==  "significant")  %>% as.data.frame()
#report_file_metab_sb$feature <- rownames(report_file_metab_sb)
write.csv(report_file_metab_sb, "limma_metab_table_sb_only_age.csv", row.names = F)
report_file_metab_sb <- read.csv("limma_metab_table_sb_only_age.csv")

#plot heatmap for the top metabolites with age
# Extract top metabolites
top_metabs <- report_file_metab_sb %>%
#  tibble::rownames_to_column("feature") %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::slice(1:25) %>%
  dplyr::pull(feature)
top_metabs <- top_metabs[which(top_metabs %!in% c("Asn His Lys", "Ala Trp Arg Esi+1.0569996" , "> limit Esi-12.231005"))]
# Extract metabolite rows
heat <- metabolite_matrix_check_SB[top_metabs, ]

# Order columns (samples) by age
metadata_ord <- metadata_SB_limma %>% arrange(Age)
heat         <- heat[, c(colnames(heat) %in% metadata_ord$normalized_name)]
head(heat)
# Z-score each metabolite across samples
heat_scaled <- t(scale(t(heat)))

# Clean display names for heatmap only
clean_labels <- c("Pregnenolone sulfate",
"Allylestrenol","Allylestrenol",
"3Z,6Z,9Z,11Z-Nonadecatetraene",
"6Z,9Z,12Z-Nonadecatriene",
"3α,11β,17α-Trihydroxy-5β-pregnan-20-one",
"1-Hexanol arabinosylglucoside",
"2,5-Dihydro-2,4-dimethyloxazole",
"6-Bromoheneicosa-5E,9Z-dienoic acid",
"3-(8,11,14-Pentadecatrienyl)phenol",
"Okanin 4-methyl ether caffeoyl-acetylglucoside",
"17α,2α-Dimethyl-17β-hydroxy-4-androsten-3-one",
"Glucosylceramide GlcCer(d14:1/24:0(2OH))",
"5-Oxoavermectin 1b aglycone",
"Phosphatidylethanolamine PE(19:0/0:0)",
"Unknown metabolite (C18H30O2)",
"NKH477", "Monogalactosyldiacylglycerol MGDG(16:0/18:2)",
"Glucosylceramide GlcCer(d18:1/24:1)",
"17β-Estradiol 3-sulfate",
"Plasmenyl phosphatidylglycerol PG(P-20:0/18:1)", 
"[10]-Paradol")
clean_labels
#Replace row names only for visualization
rownames(heat_scaled) <- clean_labels

#Column annotation (samples)
annotation_col           <- data.frame(Age = metadata_ord$Age[match(colnames(heat_scaled), metadata_ord$metabol_sample_name)])
rownames(annotation_col) <- metadata_ord$normalized_name[match(colnames(heat_scaled), metadata_ord$normalized_name)]
#annotation_col <- annotation_col %>% arrange(Age) %>% as.data.frame()
dim(annotation_col)
dim(heat_scaled)
head(heat_scaled)
                                 
ann_colors <- list(Age = colorRampPalette(c("#4575B4", "#FFFFBF", "#D73027"))(100))

#Heatmap
metab_heatmap <- pheatmap(heat_scaled,
                          annotation_col = annotation_col, annotation_colors = ann_colors,
                          cluster_rows = TRUE,
                          cluster_cols = FALSE,
                          show_rownames = TRUE,
                          show_colnames = FALSE,
                          fontsize_row = 20, fontsize_col = 30, fontsize_number = 30, fontsize = 20,
                          treeheight_row = 40,
                          border_color = NA,
                          color = colorRampPalette((brewer.pal(9, "RdBu")))(100),
                          labels_row = rownames(heat_scaled), main = "Top 25 FDR-significant metabolites associated with age")


jpeg("metab_age_heatmapFig10.png" ,  width = 40, height = 18, units = "in", res = 450)
metab_heatmap
dev.off()
#Run PCA
pca_res <- prcomp(metab_pareto, center = FALSE, scale. = FALSE)
pca_df  <- data.frame(normalized_name = rownames(pca_res$x), PC1 = pca_res$x[,1], PC2 = pca_res$x[,2])

pca_df <- left_join(pca_df, metadata, by = "normalized_name")

#Step 7: Check PCA summary
summary(pca)

#Create PCA scores plot
library(cowplot)

pca_metabol1 <- pca_scores %>% filter(!is.na(Species)) %>%   ggplot(aes(x = PC1, y = PC2, color = Species)) + geom_point(size = 3) +
                               theme_pubr(base_size = 24) +
                               labs(title = " ", x = paste0("PC1 (", round(summary(pca)$importance[2,1]*100,1), "%)"), 
                                    y = paste0("PC2 (", round(summary(pca)$importance[2,2]*100,1), "%)")) + scale_color_manual(values = pal_npg(), name = NULL)

pca_df %>% filter(!is.na(Species)) %>% ggplot(aes(PC1, PC2, color = Species)) + geom_point(size = 3) +
                   theme_pubr(base_size = 24) +
                   stat_ellipse() + labs( title = "Metabolomics", x = paste0("PC1: ", round(summary(pca_res)$importance[2,1] * 100, 1), "%"), 
                                          y = paste0("PC2: ", round(summary(pca_res)$importance[2,2] * 100, 1), "%")) + scale_color_manual(values = c("yellow3","purple3"), name = NULL)


pca_metabol2 <- pca_scores %>% filter(!is.na(Species)) %>%   ggplot(aes(x = PC1, y = PC2, color = Age_Group)) +
                              geom_point(size = 3) +
                              theme_pubr(base_size = 24) + labs(title = " ", x = paste0("PC1 (", round(summary(pca)$importance[2,1]*100,1), "%)"),
                              y = paste0("PC2 (", round(summary(pca)$importance[2,2]*100,1), "%)")) + 
                              scale_color_manual(values = c("black", "skyblue" , "#009E73","#CC79A7","purple3" ,"gray", "lightgray"), name = NULL) + facet_wrap(.~Species) 

plot_grid(pca_metabol1, pca_metabol2, labels = c("A", "B"), label_size = 24)

rownames(metadata) <- metadata$normalized_name
#Compute euclidean distance (same as QIIME)
dist_mat <- dist(metab_pareto, method = "euclidean")
rownames(dist_mat) == rownames(metadata)

#reorder dist_mat same way as metadata 

# Convert dist object to matrix
dist_mat_matrix <- as.matrix(dist_mat)
rownames(dist_mat_matrix)
# Reorder rows/columns
dist_mat_matrix_reordered <- dist_mat_matrix[metadata$normalized_name, metadata$normalized_name]

# Convert back to dist object
dist_mat <- as.dist(dist_mat_matrix_reordered)
adon_meta <- adonis2(dist_mat ~ Species, data = metadata, permutations = 999, by = "margin") 
summary(adon_meta)
adon_meta$R2

disp <- betadisper(dist_mat, metadata$Species)
anova(disp)
plot(disp)

# Perform Principal Coordinates Analysis (PCoA)
pcoa <- cmdscale(dist_mat, k = 2, eig = TRUE)
# Extract coordinates
ordination_df <- data.frame(pcoa_res$points) %>%
  rownames_to_column("SampleID") %>%
  left_join(metadata, by = c("SampleID" = rownames(metadata)))

# Plot ordination
ggplot(ordination_df, aes(x = X1, y = X2, color = Group)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(x = "PCoA1", y = "PCoA2", title = "Bray–Curtis PCoA")
```


#Fig 10. GBM - metabolites
```{r}
#fix the GBM first
#do TSS for the func mat pathways
func_tss <- sweep(func_mat_gbm,1,rowSums(func_mat_gbm),"/")
#pathway of interest
pos                <- which(colnames(func_mat_gbm) %in% age_feat)#only picking the ones for age
pathway_tss <- func_tss[, pos]

pseudo       <- min(pathway_tss[pathway_tss > 0], na.rm = TRUE) / 2
pathway_log2 <- log2(pathway_tss + pseudo) #rows are samples, and columns are samples

annotations   <- read.csv("old_stuff/sbol_metabolite_list.csv", row.names = 1, check.names = FALSE)
head(annotations)
length(which(is.na(annotations$hmdb_id ))) #5180 - are there any ones here related to tryp
leftover_unannotated <- annotations %>% filter(is.na(hmdb_id)) %>% as.data.frame()
write.csv(leftover_unannotated, "unannotated_metabolites.csv")

new_gbm_metabolites <- sort(c("HMDB0000042", "HMDB0000929",
"HMDB0000738", "HMDB0002302", "HMDB0000197",
"HMDB0000684", "HMDB0000232", "HMDB0000148",
"HMDB0000112", "HMDB0000211",
"HMDB0000696", "HMDB0001185", "HMDB0000939", "HMDB0000742",
"HMDB0000208", "HMDB0000254", "HMDB0000902", "HMDB0001487", "HMDB0001389"))

old_gbm_metabolites_id <- sort(c("HMDB0000065", "HMDB0000132", "HMDB0000155", "HMDB0003424", "HMDB0003576", "HMDB0000148", "HMDB0000112", 
"HMDB0000208", "HMDB0000254" , "HMDB0000232",  "HMDB0001406",  "HMDB0001400", "HMDB0000902", "HMDB0001487", "HMDB0003555", 
"HMDB0241844", "HMDB0000211", "HMDB0001037", "HMDB0000929", "HMDB0000738", "HMDB0000197", "HMDB0002302", "HMDB0000259", "HMDB0001389", "HMDB0000684","HMDB0001185","HMDB0000696", "HMDB0000939", "HMDB0000742", "HMDB0000042", "HMDB0001208", "HMDB0000107"))

comp_names               <- annotations$Compound[which(annotations$hmdb_id %in% old_gbm_metabolites_id)] #only 11 found
gbm_metab_features       <- feature_table[,colnames(feature_table) %in% comp_names][1:128,]
metab_metadata           <- read.csv("metadata_for_metabolomics.csv", row.names = 1, check.names = FALSE)
metab_metadata$Age_Group     <- shotgun_metadata$Age_group1[match(metab_metadata$Animal, shotgun_metadata$Animal)]
metab_metadata$Location_Type <- shotgun_metadata$Location_Type[match(metab_metadata$Animal, shotgun_metadata$Animal)]
metab_metadata$Location_Type <- shotgun_metadata$Age[match(metab_metadata$Animal, shotgun_metadata$Animal)]

gbm_metab_features$animal    <- metab_metadata$Animal[match(rownames(gbm_metab_features), metab_metadata$normalized_name)]
gbm_metab_features$sample    <- shotgun_metadata$Sample[match(gbm_metab_features$animal, shotgun_metadata$Animal)]
gbm_metab_features <- gbm_metab_features %>% filter(!is.na(sample)) %>%  as.data.frame()
rownames(gbm_metab_features) <- gbm_metab_features$sample

gbm_metab_features <- gbm_metab_features %>% dplyr::select(-animal, -sample) %>% as.data.frame()
head(gbm_metab_features) #rownames are samples 

head(pathway_log2)
gbm_metab_features <- gbm_metab_features[match(rownames(pathway_log2),rownames(gbm_metab_features)),]

gbm_metab_features <- gbm_metab_features %>% mutate(across(where(is.character), as.numeric)) %>% as.data.frame()

pathway_df    <- pathway_log2
metabolite_df <- gbm_metab_features
head(metab_metadata)
results <- expand.grid(
  pathway = colnames(pathway_df),
  metabolite = colnames(metabolite_df),
  stringsAsFactors = FALSE
)
head(results)
results$rho <- NA
results$pval <- NA

for(i in seq_len(nrow(results))) {

  p <- pathway_df[,results$pathway[i]]
  m <- metabolite_df[,results$metabolite[i]]
  ct <- cor.test(p, m, method = "spearman", exact = FALSE)
  results$rho[i] <- unname(ct$estimate)
  results$pval[i] <- ct$p.value
}

# FDR correction across all 140 tests
results$qval <- p.adjust(results$pval, method = "fdr")

results <- results %>%
  arrange(qval)

results

sig_results <- results %>%
  filter(qval < 0.05)

sig_results
library(ggplot2)
 

results$label <- ifelse(results$qval < 0.05, "*",  "")

gbm_metab_assoc <- ggplot(results, aes( x = metabolite,y = pathway, fill = rho)) +
  geom_tile() + geom_text(aes(label = label), size = 10, color = "white") + ylab("gut brain module") + 
  scale_fill_gradient2(low = "blue", mid = "white",high = "coral",midpoint = 0 ) + theme_pubr(base_size = 24) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(legend.position = "right")
gbm_metab_assoc 
ggsave("Fig11.jpeg", gbm_metab_assoc, width = 24, height = 20, dpi = 450, units = "in") #overall correaltion

#covariate adjusted correlations
path_res <- residuals(lm(pathway ~ age + sex, data = meta))

met_res <- residuals(lm(metabolite ~ age + sex, data = meta))

cor.test(
  path_res,
  met_res,
  method = "spearman")
```


```{r}
        
#Step 1: Residualize all pathways
class(pathway_df)
pathway_df <- as.data.frame(pathway_df)
head(pathway_df)
#add metadata 
metab_metadata$samplename <- shotgun_metadata$Sample[match(metab_metadata$Animal, shotgun_metadata$Animal)]
pathway_df$age            <- metab_metadata$Age[match(rownames(pathway_df), metab_metadata$samplename)]
pathway_df$sex            <- metab_metadata$Sex[match(rownames(pathway_df), metab_metadata$samplename)]
pathway_df$species        <- metab_metadata$Species[match(rownames(pathway_df), metab_metadata$samplename)]

pathway_df_complete <- na.omit(pathway_df)
res_df <- matrix(nrow = nrow(pathway_df_complete), ncol = ncol(pathway_df_complete))
colnames(res_df) <- colnames(pathway_df_complete)
rownames(res_df) <- rownames(pathway_df_complete)
i <- 1
for(i in 1:ncol(pathway_df_complete)){
fit <-   lm(pathway_df_complete[,i] ~ species + age + sex, data = pathway_df_complete)
#Extract residuals and store them safely
  res_df[, i] <- resid(fit)
    }

# 3. Convert back to a clean dataframe if needed
res_df <- as.data.frame(res_df)

#Step 2: Residualize all metabolites

#Step 3: Correlate all residualized pathways against all residualized metabolites
results <- expand.grid(
  pathway = colnames(res_df),
  metabolite = colnames(metabolite_resid),
  stringsAsFactors = FALSE
)

results$rho <- NA
results$pval <- NA

for(i in seq_len(nrow(results))) {

  p <- pathway_resid[[results$pathway[i]]]
  m <- metabolite_resid[[results$metabolite[i]]]

  ct <- cor.test(
    p,
    m,
    method = "spearman",
    exact = FALSE
  )

  results$rho[i] <- unname(ct$estimate)
  results$pval[i] <- ct$p.value
}

results$qval <- p.adjust(
  results$pval,
  method = "fdr")

results <- results[order(results$qval),]

head(results)
Significant hits
sig_results <- subset(results, qval < 0.05)

sig_results
Scatterplot for a significant pair
Using the residualized values:
path <- sig_results$pathway[1]
met  <- sig_results$metabolite[1]

plot_df <- data.frame(
  pathway = pathway_resid[[path]],
  metabolite = metabolite_resid[[met]]
)

library(ggpubr)

ggscatter(
  plot_df,
  x = "pathway",
  y = "metabolite",
  add = "reg.line",
  conf.int = TRUE
) +
stat_cor(method = "spearman") +
labs(
  x = paste(path, "(age/sex adjusted)"),
  y = paste(met, "(age/sex adjusted)")
)
One additional thought: if your primary pathway models were
pathway ~ species + age + sex
and you want pathway–metabolite associations independent of disease group as well, you may want to residualize on:
~ species + age + sex

```

#Supplementary section figures for BMC Bio

#Fig S1 - taxa barplot (moved this to the supp section for now)
```{r}
#Merge all ASVs/OTUs belonging to the same phylum.
ps_phylum    <- tax_glom(ps_rel, taxrank = "Phylum")
phylum_table <- otu_table(ps_phylum)
phylum_table <- as.data.frame(phylum_table)

names(sort(taxa_sums(ps_rel), decreasing = TRUE))

#remove NAs or unknown phylum
tax_table(ps_phylum)[, "Phylum"][is.na(tax_table(ps_phylum)[, "Phylum"])] <- "Unknown"
df_phylum <- psmelt(ps_phylum)
df_phylum <- df_phylum %>% filter(Abundance > 0) %>% as.data.frame() # REMOVE 0s HERE

num_phyla <- df_phylum %>% 
  group_by(sample.id) %>%
  dplyr::summarize(total_phyla = length(unique(Phylum))) 
num_phyla$total_phyla

top_phyla        <- names(sort(tapply(df_phylum$Abundance, df_phylum$Phylum, sum), decreasing = TRUE))[1:10]
df_phylum$Phylum <- ifelse(df_phylum$Phylum %in% top_phyla,df_phylum$Phylum, "Other")

#Keep only relevant taxonomy columns
colnames(df_phylum) #clean this up! 
df_phylum <- df_phylum %>% dplyr::select("sample.id"  ,"Phylum", "Species", "Sex", "Age", "Location_type", "is.neg", "Abundance" ) %>% as.data.frame()
head(df_phylum)

#renaming species for plots
df_phylum$Species <- ifelse(df_phylum$Species == "SB", "boliviensis", "sciureus")

df_phyla_top_10 <- df_phylum %>%
  group_by(Phylum, Species) %>%
  dplyr::summarize(total_abundance = sum(Abundance)) %>% ungroup() %>% group_by(Species) %>% mutate(all_total = sum(total_abundance)) %>% 
  mutate(percent = 100 * (total_abundance / all_total)) %>%
  arrange(desc(percent))

#Do the same for family level plot
ps_family    <- tax_glom(ps_rel, taxrank = "Family")
family_table <- otu_table(ps_family)
family_table <- as.data.frame(family_table)

#remove NAs or unknown phylum
tax_table(ps_family)[, "Family"][is.na(tax_table(ps_family)[, "Family"])] <- "Unknown"
df_family <- psmelt(ps_family)
df_family <- df_family %>% filter(Abundance >0) %>% as.data.frame()

num_family <- df_family %>% 
  group_by(sample.id) %>%
  dplyr::summarize(total_family = length(unique(Family))) 

num_family$total_family

top_family <- names(sort(tapply(df_family$Abundance, df_family$Family, sum), decreasing = TRUE))[1:25]

df_family$Family <- ifelse(df_family$Family %in% top_family,df_family$Family, "Other")

#Keep only relevant taxonomy columns
colnames(df_family) #clean this up! 
df_family <- df_family %>% dplyr::select("sample.id"  ,"Family", "Species", "Sex", "Age", "Location_type", "is.neg", "Abundance" ) %>% as.data.frame()

df_family$Species <- ifelse(df_family$Species == "SB", "boliviensis", "sciureus")

df_family_top <- df_family %>%
  group_by(Family, Species) %>%
  dplyr::summarize(total_abundance = sum(Abundance)) %>% ungroup() %>% group_by(Species) %>% mutate(all_total = sum(total_abundance)) %>% 
  mutate(percent = 100 * (total_abundance / all_total)) %>%
  arrange(desc(percent))

#do the same for geneus level too
ps_genus <- tax_glom(ps_rel, taxrank = "Genus")

tax_table(ps_genus)[, "Genus"][is.na(tax_table(ps_genus)[, "Genus"])] <- "Unknown"

df_genus <- psmelt(ps_genus) %>%
  filter(Abundance > 0) %>%
  as.data.frame()

top_genus <- names(sort(tapply(df_genus$Abundance, df_genus$Genus, sum), decreasing = TRUE))[1:25]

df_genus$Genus <- ifelse(df_genus$Genus %in% top_genus, df_genus$Genus, "Other")

df_genus <- df_genus %>%
  dplyr::select("sample.id","Genus","Species","Sex","Age","Location_type","is.neg","Abundance")

df_genus$Species <- ifelse(df_genus$Species == "SB", "boliviensis", "sciureus")
sample_order     <- unique(df_phylum$sample.id)

df_phylum$sample.id  <- factor(df_phylum$sample.id, levels = sample_order)
df_family$sample.id  <- factor(df_family$sample.id, levels = sample_order)
df_genus$sample.id   <- factor(df_genus$sample.id, levels = sample_order)

#add plot
##set the theme + plot
t1 <- theme(                               
  plot.background = element_blank(), 
  panel.grid.major = element_blank(), 
  panel.grid.minor = element_blank(), 
  panel.border = element_blank(), 
  panel.background = element_blank(),
  axis.line.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.line.y = element_line(size=.4), 
  legend.position= "right",
  legend.title = element_blank(),
  legend.text= element_text(size=14),
  axis.title.x = element_text(color="black", size=18),
  axis.title.y = element_text(color="black", size=18),
  axis.text.y = element_text(color="black", size=18),
  axis.text.x = element_blank())

#use color palette from https://sashat.me/2017/01/11/list-of-20-simple-distinct-colors/
phy_cols <- c('#3cb44b', '#ffe119', '#4363d8', '#f58231', '#bfef45', '#911eb4', '#42d4f4', '#f032e6',  '#fabebe', '#469990', "black") #can change this

Figure_S1A <- ggplot(df_phylum, aes(x= sample.id, y = Abundance, fill = Phylum)) + 
  geom_bar(stat="identity") +
    scale_fill_manual(values = phy_cols) +
  ylab("Relative abundance of phylum") +
  xlab(" ") +  t1 + facet_grid(.~Species, scales = "free_x") + theme(strip.text = element_text(face = "italic", size = 20))
Figure_S1A

fam_cols <- scales::hue_pal()(length(unique(df_family$Family)))  # #get color palette for top 20 family

Figure_S1B <- ggplot(df_family, aes(x= sample.id, y = Abundance, fill = Family)) + 
  geom_bar(stat = "identity") +
    scale_fill_manual(values = fam_cols) +
  ylab("Relative abundance of family") +
  xlab(" ") +  t1 + facet_grid(.~Species, scales = "free_x") + theme(strip.text = element_text(face = "italic", size = 20)) 
Figure_S1B

gen_cols        <- scales::hue_pal()(length(unique(df_genus$Genus)))  # simple palette

Figure_S1C <- ggplot(df_genus, aes(x = sample.id, y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = gen_cols) +
  ylab("Relative abundance of genus") +
  xlab("Sample") +
  t1 + facet_grid(. ~ Species, scales = "free_x") + theme(strip.text = element_text(size = 20))

Figure_S1C

final_Fig1 <- Figure_S1A / Figure_S1B  +
  plot_layout(nrow = 2, heights = c(1,1)) + plot_annotation(tag_levels = 'A')

final_Fig1 #doesn't look good
Figure_1B <- df_genus_summ %>% ggplot(aes(x = sample.id, y = Sum_abund, fill=Genus)) + 
  geom_bar(stat="identity") +
    scale_fill_manual(values = genus_cols) +
  ylab("Relative abundance of genus") +
  xlab("Sample ID") +  facet_grid(.~Species, scales = "free_x") + theme(strip.text = element_text(size = 20)) 
Figure_1B 
#save this 
ggsave("BMC_Bio_FigS1.png", final_Fig1, width = 24, height = 18, units = "in", dpi = 400 )

#t-test 
wilcox.test(df_genus_summ$Sum_abund[df_genus_summ$genus_cat ==  "Butyrate producers"] ~ df_genus_summ$Species[df_genus_summ$genus_cat ==  "Butyrate producers"] )
effectsize::rank_biserial(df_genus_summ$Sum_abund[df_genus_summ$genus_cat ==  "Butyrate producers"] ~ df_genus_summ$Species[df_genus_summ$genus_cat ==  "Butyrate producers"])  #sig

#check for t-test 
summary(lm(df_genus_summ$Sum_abund[df_genus_summ$genus_cat ==  "Lactate producers"] ~ df_genus_summ$Species[df_genus_summ$genus_cat ==  "Lactate producers"] ))
effectsize::rank_biserial(df_genus_summ$Sum_abund[df_genus_summ$genus_cat ==  "Lactate producers"] ~ df_genus_summ$Species[df_genus_summ$genus_cat ==  "Lactate producers"]) #sig

#check for t-test 
wilcox.test(df_genus_summ$Sum_abund[df_genus_summ$genus_cat ==  "Carbohydrate degraders"] ~ df_genus_summ$Species[df_genus_summ$genus_cat ==  "Carbohydrate degraders"] )
effectsize::rank_biserial(df_genus_summ$Sum_abund[df_genus_summ$genus_cat ==  "Carbohydrate degraders"] ~ df_genus_summ$Species[df_genus_summ$genus_cat ==  "Carbohydrate degraders"]) #ns

Supp6c.1 <- df_genus_summ %>% filter(genus_cat == "Butyrate producers" & !is.na(Species)) %>% ggplot(aes(x = Species, y = log1p(Sum_abund + 1e-6), color = Species)) + geom_violin() + geom_jitter(size = 3, alpha = 0.62) + theme_pubr(base_size = 24) + scale_color_bmj() + xlab("") + ylab("Relative abundance") +  ggtitle("butyrate producing taxa") + geom_signif(comparisons = list(c("SB", "SC")), map_signif_level = TRUE, color = "black",  test = "wilcox.test", step_increase = 0.1 ) + theme(axis.text.x = element_blank())

Supp6c.2 <- df_genus_summ %>% filter(genus_cat == "Lactate producers" & !is.na(Species)) %>% ggplot(aes(x = Species, y = log1p(Sum_abund + 1e-6), color = Species)) + geom_violin() + geom_jitter(size = 3, alpha = 0.62) + theme_pubr(base_size = 24) + scale_color_bmj() + xlab("") + ylab(" ") +  ggtitle("lactate producing taxa") + geom_signif(comparisons = list(c("SB", "SC")), map_signif_level = TRUE, color = "black",  test = "wilcox.test", step_increase = 0.1 ) + theme(axis.text.x = element_blank())

Supp6c.3 <- df_genus_summ %>% filter(genus_cat == "Carbohydrate degraders" & !is.na(Species)) %>% ggplot(aes(x = Species, y = log1p(Sum_abund + 1e-6), color = Species)) + geom_violin() + geom_jitter(size = 3, alpha = 0.62) + theme_pubr(base_size = 24) + scale_color_bmj() + xlab("") + ylab(" ") +  ggtitle("carbohydrate degrading taxa") + geom_signif(comparisons = list(c("SB", "SC")), map_signif_level = TRUE, color = "black",  test = "wilcox.test", step_increase = 0.1 ) + theme(axis.text.x = element_blank())
unique(df_genus_summ$Genus)

benefic_microbes_species <- plot_grid(Supp6c.1, Supp6c.2, Supp6c.3, ncol = 3)
benefic_microbes_species
ggsave("fig2c_subplot_benefic_microbes_Species.jpeg",benefic_microbes,  height = 6, width= 18, dpi = 350)


cor.test(df_genus_summ$Sum_abund[df_genus_summ$Genus == "Faecalibacterium"], df_genus_summ$butyrate[df_genus_summ$Genus == "Faecalibacterium"], method = "spearman")
summary(lm(df_genus_summ$Sum_abund[df_genus_summ$Genus == "Faecalibacterium"] ~ df_genus_summ$Species[df_genus_summ$Genus == "Faecalibacterium"] + df_genus_summ$Age[df_genus_summ$Genus == "Faecalibacterium"])) #ns

df_genus_summ %>% filter(Genus == "Faecalibacterium") %>% ggplot(aes(y = Sum_abund, x = Age, color = Species)) + geom_point(size = 2, alpha = 0.62) + geom_smooth(method = "lm")  + scale_color_bmj() + xlab("Age") + ylab("Rel abundance") + labs(title = "Faecalibacterium") + theme_pubr(base_size = 24) + facet_grid(.~Species) + theme(legend.position = "none")

summary(lm(df_genus_summ$Sum_abund[df_genus_summ$Genus == "Agathobaculum"] ~ df_genus_summ$Species[df_genus_summ$Genus == "Agathobaculum"])) #sig
gen1 <- df_genus_summ %>% filter(Genus == "Agathobaculum" ) %>% ggplot(aes(y = Sum_abund,  x = Age, color = Species)) + geom_point(size = 2, alpha = 0.62) + geom_smooth(method = "lm")  + scale_color_bmj() + xlab("Age") + ylab("Rel abundance") + labs(title = "Agathobaculum") + theme_pubr(base_size = 24) + facet_grid(.~Species)  + theme(legend.position = "none") #significant
gen1

summary(lm(df_genus_summ$Sum_abund[df_genus_summ$Genus == "Prevotella"] ~ df_genus_summ$Species[df_genus_summ$Genus == "Prevotella"])) #ns

gen2 <- df_genus_summ %>% filter(Genus == "Prevotella" ) %>% ggplot(aes(y = Sum_abund, x = Age, color = Species)) + geom_point(size = 2, alpha = 0.62) + geom_smooth(method = "lm")  + scale_color_bmj() + xlab("Age") + ylab("Rel abundance") + labs(title = "Prevotella") + theme_pubr(base_size = 24) + facet_grid(.~Species) + theme(legend.position = "none")

summary(lm(df_genus_summ$Sum_abund[df_genus_summ$Genus == "Ruminococcus_B"] ~ df_genus_summ$Species[df_genus_summ$Genus == "Ruminococcus_B"])) #ns

gen3 <- df_genus_summ %>% filter(Genus == "Ruminococcus_B" ) %>% ggplot(aes(y = Sum_abund, x = Age, color = Species)) + geom_point(size = 2, alpha = 0.62) + geom_smooth(method = "lm")  + scale_color_bmj() + xlab("Age") + ylab("Rel abundance") + labs(title = "Ruminococcus_B") + theme_pubr(base_size = 24) + facet_grid(.~Species) + theme(legend.position = "none") #ns

summary(lm(df_genus_summ$Sum_abund[df_genus_summ$Genus == "Bifidobacterium_388775"] ~ df_genus_summ$Species[df_genus_summ$Genus == "Bifidobacterium_388775"]))

gen4 <- df_genus_summ %>% filter(Genus == "Bifidobacterium_388775" ) %>% ggplot(aes(y = Sum_abund, x = Age, color = Species)) + geom_point(size = 2, alpha = 0.62) + geom_smooth(method = "lm")  + scale_color_bmj() + xlab("Age") + ylab("Rel abundance ") + labs(title = "Bifidobacterium_388775") + theme_pubr(base_size = 24) + facet_grid(.~Species) + theme(legend.position = "none") #ns

summary(lm(df_genus_summ$Sum_abund[df_genus_summ$Genus  %in% c("Blautia_A_141780","Blautia_A_141781")] ~ df_genus_summ$Species[df_genus_summ$Genus  %in% c("Blautia_A_141780","Blautia_A_141781")]))

gen5 <- df_genus_summ %>% filter(Genus %in% c("Blautia_A_141780","Blautia_A_141781" )) %>% ggplot(aes(y = Sum_abund, x = Age, color = Species)) + geom_point(size = 2, alpha = 0.62) + geom_smooth(method = "lm")  + scale_color_bmj() + xlab("Age") + ylab("Rel abundance") +  labs(title = "Blautia_A") + theme_pubr(base_size = 24) + facet_grid(.~Species) + theme(legend.position = "none") #significant

summary(lm(df_genus_summ$Sum_abund[df_genus_summ$Genus  == "Fusicatenibacter"] ~ df_genus_summ$Species[df_genus_summ$Genus  == "Fusicatenibacter"])) #ns

gen6 <- df_genus_summ %>% filter(Genus == "Fusicatenibacter" ) %>% ggplot(aes(y = Sum_abund, x = Age, color = Species)) + geom_point(size = 2, alpha = 0.62) + geom_smooth(method = "lm")  + scale_color_bmj() + xlab("Age") + ylab("Rel abundance") + labs(title = "Fusicatenibacter") + theme_pubr(base_size = 24) + facet_grid(.~Species) + theme(legend.position = "none")

gen1 / gen5
```
#Fig S2  - are there any microbial taxa unique to each species
```{r}
#Looking for genus level prevalences 
ps_genus
range(sample_sums(ps_genus))

# Extract data
otu  <- as.data.frame(otu_table(ps_genus))
meta <- data.frame(sample_data(ps_genus))

#Convert to presence/absence
threshold                   <- 0.001
otu_pa                      <- otu
otu_pa[otu_pa < threshold]  <- 0
otu_pa[otu_pa >= threshold] <- 1

#Calculate prevalence per species

#Split samples by species
species_list <- split(meta$sample.id, meta$Species)

#Function to calculate prevalence
calc_prev <- function(samples) {
  rowMeans(otu_pa[, samples])
}

prev_bol <- calc_prev(species_list[["SB"]])
prev_sci <- calc_prev(species_list[["SC"]])

#Identify species-specific taxa
#Define:
#* Present in ≥30% of one species
#* Present in ≤5% of the other

bol_unique <- names(prev_bol[
  prev_bol >= 0.3 & prev_sci <= 0.05
])

sci_unique <- names(prev_sci[
  prev_sci >= 0.3 & prev_bol <= 0.05
])


#Inspect results
tax_table(ps_genus)[bol_unique, ]
tax_table(ps_genus)[sci_unique, ]


#Visualize (recommended for paper)
#Prevalence plot

prev_df <- data.frame(
  Genus = rownames(otu_pa),
  boliviensis = prev_bol,
  sciureus = prev_sci)

head(prev_df)
prev_df %>%
  arrange(desc(abs(boliviensis - sciureus)))

present_SB <- sum(otu_pa[taxon, species_list[["SB"]]])
absent_SB  <- length(species_list[["SB"]]) - present_SB

present_SC <- sum(otu_pa[taxon, species_list[["SC"]]])
absent_SC  <- length(species_list[["SC"]]) - present_SC

fisher.test(matrix(
  c(present_SB, absent_SB, present_SC,absent_SC),
  nrow=2))

prev_df$diff <- prev_df$boliviensis - prev_df$sciureus

prev_df %>% filter(diff != 0) %>% ggplot(aes(x = reorder(Genus, abs(diff)),
           y = diff)) + theme_pubr(base_size = 8) + theme(axis.title.y = element_blank()) +
  geom_col() + xlab("") + ylab("Differences in prevalence") +
  coord_flip()


fisher.test(
  matrix(c(
    present_SB,
    absent_SB,
    present_SC,
    absent_SC
  ), nrow = 2)
)
for (tax in selected) {
  sb_present <- sum(
    otu_pa[tax, species_list[["SB"]]]
  )
  sc_present <- sum(
    otu_pa[tax, species_list[["SC"]]]
  )
  cat(tax,
    " SB:", sb_present, "/", length(species_list[["SB"]]),
    " SC:", sc_present, "/", length(species_list[["SC"]]),
    "\n"
  )
}

taxonomy(ps)

ggplot(prev_df, aes(x= boliviensis, y= sciureus)) +
  geom_point(alpha=0.5) +
  geom_abline(slope=1, intercept=0, linetype="dashed") +
  theme_pubr(base_size = 16) + 
  xlab("Prevalence (boliviensis)") +
  ylab("Prevalence (sciureus)")



#Points far from diagonal = species-biased taxa

#Heatmap of species-specific taxa

library(pheatmap)

selected <- unique(c(bol_unique, sci_unique))
mat <- otu[selected, ]
annotation_col <- data.frame(
  Species = meta$Species
)

rownames(annotation_col) <- colnames(mat)

pheatmap(
  mat,
  scale = "row",
  annotation_col = annotation_col)
pheatmap(mat, scale="row")


#How to report this (important wording)
# Avoid: “taxa unique to species”
# Use: “taxa enriched or prevalent in one species relative to the other”



#check for ASV level differenes, 
```


#Fig S3 - age group box plots
```{r}

shotgun_metadata$Age_group1 <- factor(shotgun_metadata$Age_group1, levels = c("Infant", "Yearling", "Juvenile", "Adult", "Geriatric", "Geriatric_old"))
 
shotgun_metadata$Species <- as.factor(shotgun_metadata$Species)
shotgun_metadata$enclosure <- as.factor(shotgun_metadata$enclosure)

#age plots for 16S shannon - this needs fixing 
fig2a.1 <- shotgun_metadata %>% ggplot( aes(x = Age_group1, y = otu_shannon, color = Age_group1)) +
  geom_boxplot() + geom_point(alpha = 0.3, size = 2) + xlab("") + ylab("Shannon diversity \n of ASVs") + theme_pubr(base_size = 24) + 
  scale_color_manual(values = age_palette, name = NULL) +
  labs(title = " ") + theme(axis.text.x = element_blank()) + facet_grid(.~Species)

fig2a.2 <-  shotgun_metadata %>% ggplot( aes(x = Age_group1, y = otu_simpson, color = Age_group1)) +
  geom_boxplot() + geom_point(alpha = 0.3, size = 2) + xlab("") + ylab("Simpson diversity \n of functional pathways") +
  theme_pubr(base_size = 24) + scale_color_manual(values = age_palette, name = NULL) +
  labs(title = " ") + theme(axis.text.x = element_blank()) + facet_grid(.~Species)

fig2a.3 <- shotgun_metadata %>% ggplot( aes(x = Age_group1, y = otu_shannon, color = Age_group1)) +
  geom_boxplot() + geom_point(alpha = 0.3, size = 2) + xlab("") + ylab("Number of \n functional pathways") +
  theme_pubr(base_size = 24) + scale_color_manual(values = age_palette, name = NULL) +
  labs(title = " ") + theme(axis.text.x = element_blank()) + facet_grid(.~Species)

blockS1 <- plot_grid(fig2a.1, fig2a.2, fig2a.3, ncol = 3, nrow = 1)
#getting plots for shotgun div metrics

fig2b.1 <- shotgun_metadata %>% ggplot( aes(x = Age_group1, y = shannon, color = Age_group1)) +
  geom_boxplot() + geom_point(alpha = 0.3, size = 2) + xlab("") + ylab("Shannon diversity \n of functional pathways") + theme_pubr(base_size = 24) + 
  scale_color_manual(values = age_palette, name = NULL) +
  labs(title = " ") + theme(axis.text.x = element_blank()) + facet_grid(.~Species)

fig2b.2 <-  shotgun_metadata %>% ggplot( aes(x = Age_group1, y = simpson, color = Age_group1)) +
  geom_boxplot() + geom_point(alpha = 0.3, size = 2) + xlab("") + ylab("Simpson diversity \n of functional pathways") +
  theme_pubr(base_size = 24) + scale_color_manual(values = age_palette, name = NULL) +
  labs(title = " ") + theme(axis.text.x = element_blank()) + facet_grid(.~Species)

fig2b.3 <- shotgun_metadata %>% ggplot( aes(x = Age_group1, y = observed_pathways, color = Age_group1)) +
  geom_boxplot() + geom_point(alpha = 0.3, size = 2) + xlab("") + ylab("Number of \n functional pathways") +
  theme_pubr(base_size = 24) + scale_color_manual(values = age_palette, name = NULL) +
  labs(title = " ") + theme(axis.text.x = element_blank()) + facet_grid(.~Species)

blockS2 <- plot_grid(fig2b.1, fig2b.2, fig2b.3, ncol = 3, nrow = 1)


#this is based on enclosure type
enclosure_cols <- c("gray", "lightgreen", "forestgreen", "darkgreen")
shotgun_metadata$enclosure <- factor(shotgun_metadata$enclosure , levels = c("Individual", "Small", "Medium", "Large"))

enc1 <- shotgun_metadata %>% ggplot( aes(x = enclosure, y = otu_shannon, color = enclosure)) +
  geom_boxplot() + geom_point(alpha = 0.3, size =2) + xlab("") + ylab("Shannon diversity \n (ASVs)") +
  theme_pubr(base_size = 24) + scale_color_manual(values = enclosure_cols, name = NULL) +
  labs(title = " ") + theme(axis.text.x = element_blank()) + 
  geom_signif(comparisons = list(c("Large", "Small")), map_signif_level = TRUE, color = "black", test = "wilcox.test", step_increase = 0.1 ) + facet_grid(.~Species)


enc2 <- shotgun_metadata %>% ggplot( aes(x = enclosure, y = otu_richness, color = enclosure)) +
  geom_boxplot() + geom_point(alpha = 0.3, size =2) + xlab("") + ylab("Richness (ASVs)") +
  theme_pubr(base_size = 24) + scale_color_manual(values = enclosure_cols, name = NULL) +
  labs(title = " ") + theme(axis.text.x = element_blank()) + 
  geom_signif(comparisons = list(c("Large", "Small")), map_signif_level = TRUE, color = "black", test = "wilcox.test", step_increase = 0.1 ) + facet_grid(.~Species)


enc3 <- shotgun_metadata %>% ggplot( aes(x = enclosure, y = shannon, color = enclosure)) +
  geom_boxplot() + geom_point(alpha = 0.3, size =2) + xlab("") + ylab("Shannon diversity \n (functional pathways)") +
  theme_pubr(base_size = 24) + scale_color_manual(values = enclosure_cols, name = NULL) +
  labs(title = " ") + theme(axis.text.x = element_blank()) + 
  geom_signif(comparisons = list(c("Large", "Small")), map_signif_level = TRUE, color = "black", test = "wilcox.test", step_increase = 0.1 ) + facet_grid(.~Species)


enc4 <- shotgun_metadata %>% ggplot( aes(x = enclosure, y = observed_pathways, color = enclosure)) +
  geom_boxplot() + geom_point(alpha = 0.3, size =2) + xlab("") + ylab("Number of observed \n (functional pathways)") +
  theme_pubr(base_size = 24) + scale_color_manual(values = enclosure_cols, name = NULL) +
  labs(title = " ") + theme(axis.text.x = element_blank()) + 
  geom_signif(comparisons = list(c("Large", "Small")), map_signif_level = TRUE, color = "black", test = "wilcox.test", step_increase = 0.1 ) + facet_grid(.~Species)

plot_grid(enc1, enc2) 
plot_grid(enc3, enc4)

shotgun_metadata$Location_Type
loc_cols <- c("pink", "pink2", "pink4", "brown4")

loc1 <- shotgun_metadata %>% ggplot( aes(x = Location_Type, y = otu_shannon, color = Location_Type)) +
  geom_boxplot() + geom_point(alpha = 0.3, size =2) + xlab("") + ylab("Shannon diversity \n (ASVs)") +
  theme_pubr(base_size = 24)  + scale_color_manual(values = loc_cols, name = NULL) +
  labs(title = " ") + theme(axis.text.x = element_blank()) + 
  geom_signif(comparisons = list(c("Naïve social group", "Experimental group")), map_signif_level = TRUE, color = "black", test = "wilcox.test", step_increase = 0.1 )  + facet_grid(.~Species) 

loc2 <- shotgun_metadata %>% ggplot( aes(x = Location_Type, y = otu_richness, color = Location_Type)) +
  geom_boxplot() + geom_point(alpha = 0.3, size =2) + xlab("") + ylab("Richness (ASVs)") +
  theme_pubr(base_size = 24)  + scale_color_manual(values = loc_cols, name = NULL) +
  labs(title = " ") + theme(axis.text.x = element_blank())  + geom_signif(comparisons = list(c("Naïve social group", "Experimental group")), map_signif_level = TRUE, color = "black", test = "wilcox.test", step_increase = 0.1 ) + facet_grid(.~Species) 


loc3 <- shotgun_metadata %>% ggplot( aes(x = Location_Type, y = shannon, color = Location_Type)) +
  geom_boxplot() + geom_point(alpha = 0.3, size =2) + xlab("") + ylab("Shannon diversity \n (functional pathways)") +
  theme_pubr(base_size = 24)  + scale_color_manual(values = loc_cols, name = NULL) +
  labs(title = " ") + theme(axis.text.x = element_blank())  + geom_signif(comparisons = list(c("Naïve social group", "Experimental group")), map_signif_level = TRUE, color = "black", test = "wilcox.test", step_increase = 0.1 ) + facet_grid(.~Species) 


loc4 <- shotgun_metadata %>% ggplot( aes(x = Location_Type, y = observed_pathways, color = Location_Type)) +
  geom_boxplot() + geom_point(alpha = 0.3, size =2) + xlab("") + ylab("Number of observed \n (functional pathways)") +
  theme_pubr(base_size = 24)  + scale_color_manual(values = loc_cols, name = NULL) +
  labs(title = " ") + theme(axis.text.x = element_blank()) + geom_signif(comparisons = list(c("Naïve social group", "Experimental group")), map_signif_level = TRUE, color = "black", test = "wilcox.test", step_increase = 0.1 )   + facet_grid(.~Species) 


age_palette <- c("black", "skyblue" , "#009E73","#CC79A7","purple3" ,"gray")


nmds_df$Age_Group1 <- shotgun_metadata$Age_group1[match(nmds_df$Animal, shotgun_metadata$Animal)]

nmds_age <- nmds_df %>% filter(!is.na(Species)) %>% ggplot(aes(NMDS1, NMDS2,
                       color = Age_Group1)) +
  geom_point(size = 3, alpha = 0.5) +
  stat_ellipse(level = 0.95, geom = "polygon", alpha = 0.02) +
  facet_grid(. ~ Species) + theme_pubr(base_size = 24) +labs(title = paste("NMDS (Stress =", round(nmds_res$stress, 3), ")"),
       x = "NMDS1", y = "NMDS2") + scale_color_manual(values = age_palette,
                     name = NULL)
nmds_age
```


